---
title: "Week 6: Wednesday Solutions"
author: "Harald Kugler"
date: "10.07.2026"
---


```{r}
#| message: false
library(RwR)
library(survival)
library(ggplot2)
library(dplyr)
library(broom)
library(splines)

theme_set(theme_bw())
```

## Exercise 11.5

We consider the data set `lung` on lung cancer survival from the RwR package.

```{r}
data("lung", package = "RwR")
lung <- lung |>
    mutate(sex = factor(sex, levels = c(1, 2), labels = c("male", "female")))
head(lung)
```

The variable `status` is a factor with levels `1` (censored) and `2` (dead), so the status indicator is `status == "2"`.

### 11.5.1

*Plot the Kaplan-Meier estimates of the survival functions for males and females separately.*

As in Section 11.2.3 we use `survfit()` with `sex` in the formula to get a Kaplan-Meier estimate for each group.

```{r}
lung_KM <- survfit(Surv(time, status == "2") ~ sex, data = lung)
lung_KM
```

```{r}
summary(lung_KM, data.frame = TRUE) |>
    ggplot(aes(time, surv, color = strata)) +
    geom_step() +
    ylim(0, 1) +
    xlab("Time (days)") +
    ylab("Survival probability") +
    theme(legend.position = "top")
```

Women have better survival than men, with an estimated median survival time of 426 days compared to 270 days for men.

### 11.5.2

*Use the Cox proportional hazards model to investigate if the survival time depends upon age.*

We first fit a Cox model with `age` as the only predictor and use a likelihood-ratio test of $H_0: \beta_{\text{age}} = 0$.

```{r}
lung_cox_age <- coxph(Surv(time, status == "2") ~ age, data = lung)
lung_cox_age |> tidy()
anova(lung_cox_age)
```

The estimate is positive, so the hazard increases with age, by a factor $e^{0.0187} \approx 1.019$ per year. The test is significant ($p \approx 0.04$).

Since 11.5.1 shows that survival differs between men and women, we also test if survival depends on age given sex.

```{r}
lung_cox_sex <- coxph(Surv(time, status == "2") ~ sex, data = lung)
lung_cox <- coxph(Surv(time, status == "2") ~ age + sex, data = lung)
lung_cox |> tidy()
anova(lung_cox_sex, lung_cox)
```

Given sex, the age effect is similar in size but no longer significant at the 5% level ($p \approx 0.06$). The evidence that survival depends on age is therefore weak. The effect of sex is clear: women have a lower hazard than men.

### 11.5.3

*Check if the Cox proportional hazards model fits the data.*

We check the model `lung_cox` with both `age` and `sex` using the residuals from Section 11.5: a Cox-Snell residual plot, martingale residuals against the fitted values and against `age`, and the squared-martingale residuals for each sex (as in Example 11.12).

```{r}
#| fig-height: 8
lung_cox_aug <- augment(lung_cox, data = lung) |>
    mutate(
        Lambda = (status == "2") - .resid,  # Cox-Snell residuals
        v = .resid^2 - Lambda               # Squared-martingale residuals
    )

p1 <- augment(lung_cox, data = lung, type.predict = "expected") |>
    survfit(Surv(.fitted, status == "2") ~ 1, data = _) |>
    summary(data.frame = TRUE) |>
    ggplot(aes(time, cumhaz)) +
    geom_abline(slope = 1, color = "blue") +
    geom_point() +
    xlab("Cox-Snell residuals") +
    ylab("Cumulative hazard")

p2 <- ggplot(lung_cox_aug, aes(.fitted, .resid)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", formula = y ~ x) +
    xlab("Fitted values") +
    ylab("Martingale residuals")

p3 <- ggplot(lung_cox_aug, aes(age, .resid)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", formula = y ~ x) +
    xlab("Age") +
    ylab("Martingale residuals")

p4 <- ggplot(lung_cox_aug, aes(sex, v)) +
    geom_point(position = position_jitter(width = 0.15, seed = 1), alpha = 0.4) +
    geom_point(stat = "summary", fun = mean, color = "blue", size = 2.5) +
    geom_errorbar(
        stat = "summary",
        fun.data = mean_se,
        fun.args = list(mult = 1.96),
        color = "blue",
        width = 0.15
    ) +
    xlab("Sex") +
    ylab("Squared-martingale residuals")

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

The Cox-Snell residuals follow the line with slope 1. The smoothed martingale residuals are close to 0 against both the fitted values and age, apart from the ends where there are few observations, so there is no sign that age should enter nonlinearly. The 95% confidence intervals for the means of the squared-martingale residuals include 0 for both sexes, so there is no evidence against the proportional hazards assumption for sex.

As a formal check of the linear age effect, we test it against a natural cubic spline expansion of age.

```{r}
lung_cox_ns <- coxph(Surv(time, status == "2") ~ ns(age, df = 3) + sex, data = lung)
anova(lung_cox, lung_cox_ns)
```

The nonlinear model is not significantly better. Overall, the Cox model with a linear age effect and sex fits the data reasonably well.
