---
title: "Week 6: Wednesday Solutions"
author: "Harald Kugler"
date: "10.07.2026"
---


```{r}
#| message: false
library(RwR)
library(survival)
library(ggplot2)
library(dplyr)
library(broom)
library(splines)

theme_set(theme_bw())
```

## Exercise 11.5

We consider the data set `lung` on lung cancer survival from the RwR package.

```{r}
data("lung", package = "RwR")
lung <- lung |>
    mutate(sex = factor(sex, levels = c(1, 2), labels = c("male", "female")))
head(lung)
```

The variable `status` is a factor with levels `1` (censored) and `2` (dead), so the status indicator is `status == "2"`.

### 11.5.1

*Plot the Kaplan-Meier estimates of the survival functions for males and females separately.*

As in Section 11.2.3 we use `survfit()` with `sex` in the formula to get a Kaplan-Meier estimate for each group.

```{r}
lung_KM <- survfit(Surv(time, status == "2") ~ sex, data = lung)
lung_KM
```

```{r}
summary(lung_KM, data.frame = TRUE) |>
    ggplot(aes(time, surv, color = strata)) +
    geom_step() +
    ylim(0, 1) +
    xlab("Time (days)") +
    ylab("Survival probability") +
    theme(legend.position = "top")
```

Women have better survival than men, with an estimated median survival time of 426 days compared to 270 days for men.

### 11.5.2

*Use the Cox proportional hazards model to investigate if the survival time depends upon age.*

We first fit a Cox model with `age` as the only predictor and use a likelihood-ratio test of $H_0: \beta_{\text{age}} = 0$.

```{r}
lung_cox_age <- coxph(Surv(time, status == "2") ~ age, data = lung)
lung_cox_age |> tidy()
anova(lung_cox_age)
```

The estimate is positive, so the hazard increases with age, by a factor $e^{0.0187} \approx 1.019$ per year. The test is significant ($p \approx 0.04$).

Since 11.5.1 shows that survival differs between men and women, we also test if survival depends on age given sex.

```{r}
lung_cox_sex <- coxph(Surv(time, status == "2") ~ sex, data = lung)
lung_cox <- coxph(Surv(time, status == "2") ~ age + sex, data = lung)
lung_cox |> tidy()
anova(lung_cox_sex, lung_cox)
```

Given sex, the age effect is similar in size but no longer significant at the 5% level ($p \approx 0.06$). The evidence that survival depends on age is therefore weak. The effect of sex is clear: women have a lower hazard than men.

### 11.5.3

*Check if the Cox proportional hazards model fits the data.*

We check the model `lung_cox` with both `age` and `sex` using the residuals from Section 11.5: a Cox-Snell residual plot, martingale residuals against the fitted values and against `age`, and the squared-martingale residuals for each sex (as in Example 11.12).

```{r}
#| fig-height: 8
lung_cox_aug <- augment(lung_cox, data = lung) |>
    mutate(
        Lambda = (status == "2") - .resid,  # Cox-Snell residuals
        v = .resid^2 - Lambda               # Squared-martingale residuals
    )

p1 <- augment(lung_cox, data = lung, type.predict = "expected") |>
    survfit(Surv(.fitted, status == "2") ~ 1, data = _) |>
    summary(data.frame = TRUE) |>
    ggplot(aes(time, cumhaz)) +
    geom_abline(slope = 1, color = "blue") +
    geom_point() +
    xlab("Cox-Snell residuals") +
    ylab("Cumulative hazard")

p2 <- ggplot(lung_cox_aug, aes(.fitted, .resid)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", formula = y ~ x) +
    xlab("Fitted values") +
    ylab("Martingale residuals")

p3 <- ggplot(lung_cox_aug, aes(age, .resid)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", formula = y ~ x) +
    xlab("Age") +
    ylab("Martingale residuals")

p4 <- ggplot(lung_cox_aug, aes(sex, v)) +
    geom_point(position = position_jitter(width = 0.15, seed = 1), alpha = 0.4) +
    geom_point(stat = "summary", fun = mean, color = "blue", size = 2.5) +
    geom_errorbar(
        stat = "summary",
        fun.data = mean_se,
        fun.args = list(mult = 1.96),
        color = "blue",
        width = 0.15
    ) +
    xlab("Sex") +
    ylab("Squared-martingale residuals")

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

The Cox-Snell residuals follow the line with slope 1. The smoothed martingale residuals are close to 0 against both the fitted values and age, apart from the ends where there are few observations, so there is no sign that age should enter nonlinearly. The 95% confidence intervals for the means of the squared-martingale residuals include 0 for both sexes, so there is no evidence against the proportional hazards assumption for sex.

As a formal check of the linear age effect, we test it against a natural cubic spline expansion of age.

```{r}
lung_cox_ns <- coxph(Surv(time, status == "2") ~ ns(age, df = 3) + sex, data = lung)
anova(lung_cox, lung_cox_ns)
```

The nonlinear model is not significantly better. Overall, the Cox model with a linear age effect and sex fits the data reasonably well.



## Exercise 11.7

We consider censored survival data $(T_i, \delta_i, X_i)$, $i = 1, \ldots, n$, with $X_i \in \mathbb{R}$ and relative risks $\psi_i = e^{\beta X_i}$. The survival times are ordered without ties, $T_1 < T_2 < \ldots < T_n$.

### 11.7.1

*Show that Cox's partial log-likelihood equals*
$$
\ell(\beta) = \sum_{i: \delta_i = 1} \beta X_i - \log \Big(\sum_{j \geq i} e^{\beta X_j} \Big),
$$
*and compute the first and second derivative of $\ell$ w.r.t. $\beta$.*

By Definition 11.7, the partial likelihood is
$$
L_{\mathrm{par}} = \prod_{i: \delta_i = 1} \frac{\psi_i}{\Psi_i}, \qquad \Psi_i = \sum_{j: T_i \leq T_j} \psi_j.
$$
Since the survival times are ordered without ties, $\{j : T_i \leq T_j\} = \{j : j \geq i\}$. With $\psi_j = e^{\beta X_j}$ (11.17) we get
$$
\ell(\beta) = \log L_{\mathrm{par}} = \sum_{i: \delta_i = 1} \big( \log \psi_i - \log \Psi_i \big) = \sum_{i: \delta_i = 1} \beta X_i - \log \Big(\sum_{j \geq i} e^{\beta X_j} \Big).
$$

For the derivatives we introduce, for $k = 0, 1, 2$,
$$
\Psi_i^{(k)}(\beta) = \sum_{j \geq i} X_j^k e^{\beta X_j},
$$
so that $\Psi_i^{(0)} = \Psi_i$ and $\frac{\mathrm{d}}{\mathrm{d}\beta} \Psi_i^{(k)} = \Psi_i^{(k+1)}$. Then
$$
\ell'(\beta) = \sum_{i: \delta_i = 1} \left( X_i - \frac{\Psi_i^{(1)}(\beta)}{\Psi_i^{(0)}(\beta)} \right),
$$
and by the quotient rule
$$
\ell''(\beta) = - \sum_{i: \delta_i = 1} \left( \frac{\Psi_i^{(2)}(\beta)}{\Psi_i^{(0)}(\beta)} - \left(\frac{\Psi_i^{(1)}(\beta)}{\Psi_i^{(0)}(\beta)}\right)^2 \right).
$$

### 11.7.2

*Show that $\ell$ is concave. Find a condition on the $X_i$-s that ensures that $\ell$ has a unique maximum.*

For fixed $i$, the weights
$$
w_{ij}(\beta) = \frac{e^{\beta X_j}}{\Psi_i^{(0)}(\beta)}, \qquad j \geq i,
$$
are positive and sum to 1, so they define a probability distribution on the risk set $\{j : j \geq i\}$. With $\bar{X}_i(\beta) = \sum_{j \geq i} w_{ij}(\beta) X_j = \Psi_i^{(1)}(\beta) / \Psi_i^{(0)}(\beta)$,
$$
\frac{\Psi_i^{(2)}(\beta)}{\Psi_i^{(0)}(\beta)} - \left(\frac{\Psi_i^{(1)}(\beta)}{\Psi_i^{(0)}(\beta)}\right)^2 = \sum_{j \geq i} w_{ij}(\beta) \big(X_j - \bar{X}_i(\beta)\big)^2 \geq 0,
$$
which is the variance of the $X_j$-s in the risk set under the weights $w_{ij}(\beta)$. Hence $\ell''(\beta) \leq 0$ for all $\beta$, and $\ell$ is concave.

For the condition, we rewrite the partial log-likelihood as
$$
\ell(\beta) = - \sum_{i: \delta_i = 1} \log \Big(\sum_{j \geq i} e^{\beta (X_j - X_i)} \Big),
$$
where every term is $\leq 0$, because the summand with $j = i$ equals 1. Then $\ell$ has a unique maximum if

a) there is an uncensored $i$ and a $j > i$ with $X_j > X_i$, and
b) there is an uncensored $k$ and an $l > k$ with $X_l < X_k$.

Under a), the $i$-th term tends to $-\infty$ for $\beta \to \infty$, and since all other terms are $\leq 0$, $\ell(\beta) \to -\infty$ for $\beta \to \infty$. Likewise, b) gives $\ell(\beta) \to -\infty$ for $\beta \to -\infty$, so the continuous function $\ell$ attains a maximum. Moreover, a) implies that the $X_j$-s in the risk set of the uncensored $i$ are not all equal, so the variance above is strictly positive and $\ell''(\beta) < 0$. Thus $\ell$ is strictly concave and the maximizer is unique.

The condition is also necessary: if a) fails, $X_i \geq \bar{X}_i(\beta)$ for all uncensored $i$, so $\ell'(\beta) \geq 0$ and $\ell$ is nondecreasing without a unique maximum (similarly if b) fails). In words, the $X$-values of those who die must not be separated from the $X$-values of those still at risk.

### 11.7.3

*Implement Newton's algorithm for computing $\hat{\beta}$. Apply the algorithm with start value $\beta_1 = 0$ to the simulated data set and investigate the values of $\beta_m$, $\ell(\beta_m)$, $\ell'(\beta_m)$ and $\ell''(\beta_m)$ in the iterations of the algorithm.*

*Further: A naive implementation of the computation of e.g. $\ell$ scales like $n^2$. Make sure that your implementation scales like $n$.*

We simulate $X_i \sim N(0, 1)$, take the baseline hazard to be $\lambda_0(t) = 1$, so that $T_i^* \sim \text{Exp}(e^{\beta X_i})$, and let the censoring times be uniform on $[0, 3]$. The data are sorted by the observed times. We simulate one data set with $\beta = 1$ and one with $\beta = 0$.

```{r}
sim_cox <- function(n, beta) {
    x      <- rnorm(n)
    T_star <- rexp(n, rate = exp(beta * x))  # baseline hazard lambda_0(t) = 1
    C      <- runif(n, 0, 3)
    dat    <- data.frame(
        time   = pmin(T_star, C),
        status = as.integer(T_star <= C),
        x      = x
    )
    dat[order(dat$time), ]
}

set.seed(2026)
dat1 <- sim_cox(n = 200, beta = 1)
dat0 <- sim_cox(n = 200, beta = 0)
```

Since the data are sorted, $\Psi_i^{(k)} = \sum_{j \geq i} X_j^k e^{\beta X_j}$ for all $i$ is a reverse cumulative sum, which `rev(cumsum(rev(...)))` computes in $O(n)$ operations. Computing each of the $n$ sums from scratch would instead require $O(n^2)$ operations.

```{r}
# Partial log-likelihood and its first two derivatives.
# Assumes the data are sorted by time with no ties.
pl <- function(beta, x, status) {
    w    <- exp(beta * x)
    Psi0 <- rev(cumsum(rev(w)))
    Psi1 <- rev(cumsum(rev(x * w)))
    Psi2 <- rev(cumsum(rev(x^2 * w)))
    d    <- status == 1
    list(
        l   = sum(beta * x[d] - log(Psi0[d])),
        dl  = sum(x[d] - Psi1[d] / Psi0[d]),
        d2l = -sum(Psi2[d] / Psi0[d] - (Psi1[d] / Psi0[d])^2)
    )
}

newton_cox <- function(x, status, beta = 0, tol = 1e-10, maxit = 50) {
    trace <- vector("list", maxit)
    for (m in 1:maxit) {
        v <- pl(beta, x, status)
        trace[[m]] <- data.frame(m = m, beta = beta, l = v$l, dl = v$dl, d2l = v$d2l)
        if (abs(v$dl) < tol) break
        beta <- beta - v$dl / v$d2l  # Newton step
    }
    do.call(rbind, trace)
}
```

```{r}
newton_1 <- newton_cox(dat1$x, dat1$status)
newton_1
newton_0 <- newton_cox(dat0$x, dat0$status)
newton_0
```

In both cases the algorithm converges in a few iterations. $\ell(\beta_m)$ increases, $\ell''(\beta_m)$ is negative throughout, and $\ell'(\beta_m)$ goes to 0 very fast: once close to $\hat{\beta}$, the number of correct digits roughly doubles in each iteration (quadratic convergence). For $\beta = 0$ the start value is already close to the estimate, and $\ell''$ is nearly constant, i.e., $\ell$ is close to quadratic.

To check that the implementation scales like $n$, we time one evaluation of `pl()` for increasing $n$. The time should roughly double when $n$ doubles (which it does).

```{r}
sapply(c(1e5, 2e5, 4e5, 8e5), function(n) {
    d <- sim_cox(n, beta = 1)
    system.time(for (k in 1:10) pl(1, d$x, d$status))[["elapsed"]]
})
```

### 11.7.4

*Compute and plot an estimate of the baseline cumulative hazard function.*

We use the Breslow estimator from Section 11.4.2,
$$
\hat{\Lambda}_0(t) = \sum_{i: T_i \leq t} \frac{\delta_i}{\sum_{j \geq i} e^{\hat{\beta} X_j}},
$$
which is a cumulative sum of the $\delta_i / \Psi_i^{(0)}(\hat{\beta})$-s, since the data are sorted. We use the data set simulated with $\beta = 1$.

```{r}
beta_hat <- tail(newton_1$beta, 1)
breslow <- data.frame(
    time   = dat1$time,
    cumhaz = cumsum(dat1$status / rev(cumsum(rev(exp(beta_hat * dat1$x)))))
)

ggplot(breslow, aes(time, cumhaz)) +
    geom_step() +
    geom_abline(slope = 1, color = "blue", linetype = "dashed") +
    xlab("Time") +
    ylab("Baseline cumulative hazard")
```

The baseline corresponds to $X = 0$, and the dashed line is the true baseline cumulative hazard $\Lambda_0(t) = t$. The estimate follows it closely for small $t$ and deviates more for large $t$, where few individuals are at risk.

### 11.7.5

*Compute $\hat{\beta}$ using the `coxph()` function in R. Compute also a standard 95% confidence interval for $\beta$. Compare this interval with a (partial) likelihood interval based on $\ell$.*

```{r}
cox_fit <- coxph(Surv(time, status) ~ x, data = dat1)
c(newton = beta_hat, coxph = unname(coef(cox_fit)))
c(newton = pl(beta_hat, dat1$x, dat1$status)$l, coxph = cox_fit$loglik[2])
```

`coxph()` gives the same estimate and the same maximal partial log-likelihood as our implementation. Its baseline cumulative hazard (at $X = 0$) is also the same as our Breslow estimate:

```{r}
max(abs(basehaz(cox_fit, centered = FALSE)$hazard - breslow$cumhaz))
```

The standard 95% confidence interval is $\hat{\beta} \pm 1.96 \cdot \widehat{\mathrm{se}}$ with $\widehat{\mathrm{se}} = (-\ell''(\hat{\beta}))^{-1/2}$.

```{r}
confint(cox_fit)
```

The partial likelihood interval consists of the $\beta$-s with
$$
2 \big(\ell(\hat{\beta}) - \ell(\beta)\big) \leq q_{0.95}(\chi^2_1),
$$
and we find its endpoints with `uniroot()`.

```{r}
l_hat <- pl(beta_hat, dat1$x, dat1$status)$l
se    <- sqrt(-1 / pl(beta_hat, dat1$x, dat1$status)$d2l)  # same as coxph's std. error
comb  <- function(beta) {
    2 * (l_hat - pl(beta, dat1$x, dat1$status)$l) - qchisq(0.95, df = 1)
}
c(
    lower = uniroot(comb, c(beta_hat - 5 * se, beta_hat), tol = 1e-10)$root,
    upper = uniroot(comb, c(beta_hat, beta_hat + 5 * se), tol = 1e-10)$root
)
```

```{r}
beta_grid <- seq(beta_hat - 4 * se, beta_hat + 4 * se, length.out = 200)
data.frame(
    beta      = beta_grid,
    partial   = sapply(beta_grid, function(b) 2 * (l_hat - pl(b, dat1$x, dat1$status)$l)),
    quadratic = (beta_grid - beta_hat)^2 / se^2
) |>
    tidyr::pivot_longer(-beta, names_to = "type") |>
    ggplot(aes(beta, value, color = type)) +
    geom_line() +
    geom_hline(yintercept = qchisq(0.95, df = 1), linetype = "dashed") +
    xlab(expression(beta)) +
    ylab("")
```

The two intervals are almost identical, and both contain the true value $\beta = 1$. The partial likelihood interval is only slightly shifted to the right, because $\ell$ is very close to quadratic around $\hat{\beta}$.
