---
title: "Week 6: Wednesday Solutions"
author: "Harald Kugler"
date: "10.07.2026"
---


```{r}
#| message: false
library(RwR)
library(survival)
library(ggplot2)
library(dplyr)
library(broom)
library(splines)

theme_set(theme_bw())
```

## Exercise 11.5

We consider the data set `lung` on lung cancer survival from the RwR package.

```{r}
data("lung", package = "RwR")
lung <- lung |>
    mutate(sex = factor(sex, levels = c(1, 2), labels = c("male", "female")))
head(lung)
```

The variable `status` is a factor with levels `1` (censored) and `2` (dead), so the status indicator is `status == "2"`.

### 11.5.1

*Plot the Kaplan-Meier estimates of the survival functions for males and females separately.*

As in Section 11.2.3 we use `survfit()` with `sex` in the formula to get a Kaplan-Meier estimate for each group.

```{r}
lung_KM <- survfit(Surv(time, status == "2") ~ sex, data = lung)
lung_KM
```

```{r}
summary(lung_KM, data.frame = TRUE) |>
    ggplot(aes(time, surv, color = strata)) +
    geom_step() +
    ylim(0, 1) +
    xlab("Time (days)") +
    ylab("Survival probability") +
    theme(legend.position = "top")
```

Women have better survival than men, with an estimated median survival time of 426 days compared to 270 days for men.

### 11.5.2

*Use the Cox proportional hazards model to investigate if the survival time depends upon age.*

We first fit a Cox model with `age` as the only predictor and use a likelihood-ratio test of $H_0: \beta_{\text{age}} = 0$.

```{r}
lung_cox_age <- coxph(Surv(time, status == "2") ~ age, data = lung)
lung_cox_age |> tidy()
anova(lung_cox_age)
```

The estimate is positive, so the hazard increases with age, by a factor $e^{0.0187} \approx 1.019$ per year. The test is significant ($p \approx 0.04$).

Since 11.5.1 shows that survival differs between men and women, we also test if survival depends on age given sex.

```{r}
lung_cox_sex <- coxph(Surv(time, status == "2") ~ sex, data = lung)
lung_cox <- coxph(Surv(time, status == "2") ~ age + sex, data = lung)
lung_cox |> tidy()
anova(lung_cox_sex, lung_cox)
```

Given sex, the age effect is similar in size but no longer significant at the 5% level ($p \approx 0.06$). The evidence that survival depends on age is therefore weak. The effect of sex is clear: women have a lower hazard than men.

### 11.5.3

*Check if the Cox proportional hazards model fits the data.*

We check the model `lung_cox` with both `age` and `sex` using the residuals from Section 11.5: a Cox-Snell residual plot, martingale residuals against the fitted values and against `age`, and the squared-martingale residuals for each sex (as in Example 11.12).

```{r}
#| fig-height: 8
lung_cox_aug <- augment(lung_cox, data = lung) |>
    mutate(
        Lambda = (status == "2") - .resid,  # Cox-Snell residuals
        v = .resid^2 - Lambda               # Squared-martingale residuals
    )

p1 <- augment(lung_cox, data = lung, type.predict = "expected") |>
    survfit(Surv(.fitted, status == "2") ~ 1, data = _) |>
    summary(data.frame = TRUE) |>
    ggplot(aes(time, cumhaz)) +
    geom_abline(slope = 1, color = "blue") +
    geom_point() +
    xlab("Cox-Snell residuals") +
    ylab("Cumulative hazard")

p2 <- ggplot(lung_cox_aug, aes(.fitted, .resid)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", formula = y ~ x) +
    xlab("Fitted values") +
    ylab("Martingale residuals")

p3 <- ggplot(lung_cox_aug, aes(age, .resid)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", formula = y ~ x) +
    xlab("Age") +
    ylab("Martingale residuals")

p4 <- ggplot(lung_cox_aug, aes(sex, v)) +
    geom_point(position = position_jitter(width = 0.15, seed = 1), alpha = 0.4) +
    geom_point(stat = "summary", fun = mean, color = "blue", size = 2.5) +
    geom_errorbar(
        stat = "summary",
        fun.data = mean_se,
        fun.args = list(mult = 1.96),
        color = "blue",
        width = 0.15
    ) +
    xlab("Sex") +
    ylab("Squared-martingale residuals")

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

The Cox-Snell residuals follow the line with slope 1. The smoothed martingale residuals are close to 0 against both the fitted values and age, apart from the ends where there are few observations, so there is no sign that age should enter nonlinearly. The 95% confidence intervals for the means of the squared-martingale residuals include 0 for both sexes, so there is no evidence against the proportional hazards assumption for sex.

As a formal check of the linear age effect, we test it against a natural cubic spline expansion of age.

```{r}
lung_cox_ns <- coxph(Surv(time, status == "2") ~ ns(age, df = 3) + sex, data = lung)
anova(lung_cox, lung_cox_ns)
```

The nonlinear model is not significantly better. Overall, the Cox model with a linear age effect and sex fits the data reasonably well.



## Exercise 11.7

We consider censored survival data $(T_i, \delta_i, X_i)$, $i = 1, \ldots, n$, with $X_i \in \mathbb{R}$ and relative risks $\psi_i = e^{\beta X_i}$. The survival times are ordered without ties, $T_1 < T_2 < \ldots < T_n$.

### 11.7.1

*Show that Cox's partial log-likelihood equals*
$$
\ell(\beta) = \sum_{i: \delta_i = 1} \beta X_i - \log \Big(\sum_{j \geq i} e^{\beta X_j} \Big),
$$
*and compute the first and second derivative of $\ell$ w.r.t. $\beta$.*

By Definition 11.7, the partial likelihood is
$$
L_{\mathrm{par}} = \prod_{i: \delta_i = 1} \frac{\psi_i}{\Psi_i}, \qquad \Psi_i = \sum_{j: T_i \leq T_j} \psi_j.
$$
Since the survival times are ordered without ties, $\{j : T_i \leq T_j\} = \{j : j \geq i\}$. With $\psi_j = e^{\beta X_j}$ (11.17) we get
$$
\ell(\beta) = \log L_{\mathrm{par}} = \sum_{i: \delta_i = 1} \big( \log \psi_i - \log \Psi_i \big) = \sum_{i: \delta_i = 1} \beta X_i - \log \Big(\sum_{j \geq i} e^{\beta X_j} \Big).
$$

For the derivatives we introduce, for $k = 0, 1, 2$,
$$
\Psi_i^{(k)}(\beta) = \sum_{j \geq i} X_j^k e^{\beta X_j},
$$
so that $\Psi_i^{(0)} = \Psi_i$ and $\frac{\mathrm{d}}{\mathrm{d}\beta} \Psi_i^{(k)} = \Psi_i^{(k+1)}$. Then
$$
\ell'(\beta) = \sum_{i: \delta_i = 1} \left( X_i - \frac{\Psi_i^{(1)}(\beta)}{\Psi_i^{(0)}(\beta)} \right),
$$
and by the quotient rule
$$
\ell''(\beta) = - \sum_{i: \delta_i = 1} \frac{\Psi_i^{(0)}(\beta) \Psi_i^{(2)}(\beta) - \big( \Psi_i^{(1)}(\beta) \big)^2}{\big( \Psi_i^{(0)}(\beta) \big)^2}
= - \sum_{i: \delta_i = 1} \left( \frac{\Psi_i^{(2)}(\beta)}{\Psi_i^{(0)}(\beta)} - \left(\frac{\Psi_i^{(1)}(\beta)}{\Psi_i^{(0)}(\beta)}\right)^2 \right).
$$

### 11.7.2

*Show that $\ell$ is concave. Find a condition on the $X_i$-s that ensures that $\ell$ has a unique maximum.*

**Concavity.** 
The denominator of  $\ell''(\beta)$ is positive, and there is a minus sign in front of the sum. So $\ell''(\beta) \leq 0$ if every numerator is $\geq 0$, which is what we show now.

Fix an uncensored $i$. We show that the numerator can be written as
$$
\Psi_i^{(0)} \Psi_i^{(2)} - \big( \Psi_i^{(1)} \big)^2 = \frac{1}{2} \sum_{j \geq i} \sum_{k \geq i} e^{\beta X_j} e^{\beta X_k} \big( X_j - X_k \big)^2.
$$
To see this, start from the right-hand side and expand the square, $(X_j - X_k)^2 = X_j^2 - 2 X_j X_k + X_k^2$. Each of the resulting double sums is a product of two single sums, because a factor that only depends on $j$ can be taken outside the sum over $k$. This gives
$$
\begin{aligned}
\frac{1}{2} \sum_{j \geq i} \sum_{k \geq i} e^{\beta X_j} e^{\beta X_k} \big( X_j - X_k \big)^2
&= \frac{1}{2} \sum_{j \geq i} \sum_{k \geq i} e^{\beta X_j} e^{\beta X_k} X_j^2 - \sum_{j \geq i} \sum_{k \geq i} e^{\beta X_j} e^{\beta X_k} X_j X_k + \frac{1}{2} \sum_{j \geq i} \sum_{k \geq i} e^{\beta X_j} e^{\beta X_k} X_k^2 \\
&= \frac{1}{2} \Big( \sum_{j \geq i} X_j^2 e^{\beta X_j} \Big) \Big( \sum_{k \geq i} e^{\beta X_k} \Big) - \Big( \sum_{j \geq i} X_j e^{\beta X_j} \Big) \Big( \sum_{k \geq i} X_k e^{\beta X_k} \Big) + \frac{1}{2} \Big( \sum_{j \geq i} e^{\beta X_j} \Big) \Big( \sum_{k \geq i} X_k^2 e^{\beta X_k} \Big) \\
&= \frac{1}{2} \Psi_i^{(2)} \Psi_i^{(0)} - \big( \Psi_i^{(1)} \big)^2 + \frac{1}{2} \Psi_i^{(0)} \Psi_i^{(2)} \\
&= \Psi_i^{(0)} \Psi_i^{(2)} - \big( \Psi_i^{(1)} \big)^2.
\end{aligned}
$$
Hence
$$
\Psi_i^{(0)} \Psi_i^{(2)} - \big( \Psi_i^{(1)} \big)^2 = \frac{1}{2} \sum_{j \geq i} \sum_{k \geq i} \underbrace{e^{\beta X_j} e^{\beta X_k}}_{> 0} \underbrace{\big( X_j - X_k \big)^2}_{\geq 0} \geq 0,
$$
so $\ell''(\beta) \leq 0$ for all $\beta$, and $\ell$ is concave.

Moreover, suppose the risk set of some uncensored $i$ contains two individuals $j$ and $k$ with $X_j \neq X_k$. Then $(X_j - X_k)^2 > 0$, so that numerator is strictly positive and $\ell''(\beta) < 0$ for all $\beta$.

**Unique maximum.** Consider the condition

- **(A)** some individual dies while an individual with a larger $X$-value is still at risk, i.e., there is an uncensored $i$ and a $j > i$ with $X_j > X_i$, and
- **(B)** some individual dies while an individual with a smaller $X$-value is still at risk, i.e., there is an uncensored $k$ and an $l > k$ with $X_l < X_k$.

To see that it ensures a unique maximum, we rewrite each term of $\ell$ using $\beta X_i = \log e^{\beta X_i}$ and separate the summand with $j = i$, which equals $e^0 = 1$:
$$
\beta X_i - \log \Big( \sum_{j \geq i} e^{\beta X_j} \Big)
= - \log \Big( e^{-\beta X_i} \sum_{j \geq i} e^{\beta X_j} \Big)
= - \log \Big( \sum_{j \geq i} e^{\beta (X_j - X_i)} \Big)
= - \log \Big( 1 + \sum_{j > i} e^{\beta (X_j - X_i)} \Big).
$$
Hence
$$
\ell(\beta) = - \sum_{i: \delta_i = 1} \log \Big( \underbrace{1 + \sum_{j > i} e^{\beta (X_j - X_i)}}_{\geq 1} \Big),
$$
and every term of $\ell$ is $\leq 0$.

Existence: Under (A), the term for that $i$ contains $e^{\beta (X_j - X_i)}$ with $X_j - X_i > 0$, which tends to $\infty$ for $\beta \to \infty$. That term therefore tends to $-\infty$, and since all other terms are $\leq 0$, $\ell(\beta) \to -\infty$ for $\beta \to \infty$. In the same way, (B) gives $\ell(\beta) \to -\infty$ for $\beta \to -\infty$. A continuous function that tends to $-\infty$ in both directions attains a maximum.

Uniqueness: Under (A), the risk set of that $i$ contains the two individuals $i$ and $j$ with $X_i \neq X_j$, so $\ell''(\beta) < 0$ for all $\beta$, as shown above. A strictly concave function has at most one maximizer.


### 11.7.3

*Implement Newton's algorithm for computing $\hat{\beta}$. Apply the algorithm with start value $\beta_1 = 0$ to the simulated data set and investigate the values of $\beta_m$, $\ell(\beta_m)$, $\ell'(\beta_m)$ and $\ell''(\beta_m)$ in the iterations of the algorithm.*

*Further: A naive implementation of the computation of e.g. $\ell$ scales like $n^2$. Make sure that your implementation scales like $n$.*

We simulate $X_i \sim N(0, 1)$, take the baseline hazard to be $\lambda_0(t) = 1$, so that $T_i^* \sim \text{Exp}(e^{\beta X_i})$, and let the censoring times be uniform on $[0, 3]$. The data are sorted by the observed times. We simulate one data set with $\beta = 1$ and one with $\beta = 0$.

```{r}
sim_cox <- function(n, beta) {
    x      <- rnorm(n)
    T_star <- rexp(n, rate = exp(beta * x))  # baseline hazard lambda_0(t) = 1
    C      <- runif(n, 0, 3)
    dat    <- data.frame(
        time   = pmin(T_star, C),
        status = as.integer(T_star <= C),
        x      = x
    )
    dat[order(dat$time), ]
}

set.seed(2026)
dat1 <- sim_cox(n = 200, beta = 1)
dat0 <- sim_cox(n = 200, beta = 0)
```

Since the data are sorted, $\Psi_i^{(k)} = \sum_{j \geq i} X_j^k e^{\beta X_j}$ for all $i$ is a reverse cumulative sum, which `rev(cumsum(rev(...)))` computes in $O(n)$ operations. Computing each of the $n$ sums from scratch would instead require $O(n^2)$ operations.

```{r}
# Partial log-likelihood and its first two derivatives.
# Assumes the data are sorted by time with no ties.
pl <- function(beta, x, status) {
    w    <- exp(beta * x)
    Psi0 <- rev(cumsum(rev(w)))
    Psi1 <- rev(cumsum(rev(x * w)))
    Psi2 <- rev(cumsum(rev(x^2 * w)))
    d    <- status == 1
    list(
        l   = sum(beta * x[d] - log(Psi0[d])),
        dl  = sum(x[d] - Psi1[d] / Psi0[d]),
        d2l = -sum(Psi2[d] / Psi0[d] - (Psi1[d] / Psi0[d])^2)
    )
}

newton_cox <- function(x, status, beta = 0, tol = 1e-10, maxit = 50) {
    trace <- vector("list", maxit)
    for (m in 1:maxit) {
        v <- pl(beta, x, status)
        trace[[m]] <- data.frame(m = m, beta = beta, l = v$l, dl = v$dl, d2l = v$d2l)
        if (abs(v$dl) < tol) break
        beta <- beta - v$dl / v$d2l  # Newton step
    }
    do.call(rbind, trace)
}
```

```{r}
newton_1 <- newton_cox(dat1$x, dat1$status)
newton_1
newton_0 <- newton_cox(dat0$x, dat0$status)
newton_0
```

In both cases the algorithm converges in a few iterations. $\ell(\beta_m)$ increases, $\ell''(\beta_m)$ is negative throughout, and $\ell'(\beta_m)$ goes to 0 very fast: once close to $\hat{\beta}$, the number of correct digits roughly doubles in each iteration (quadratic convergence). For $\beta = 0$ the start value is already close to the estimate, and $\ell''$ is nearly constant, i.e., $\ell$ is close to quadratic.

To check that the implementation scales like $n$, we time one evaluation of `pl()` for increasing $n$. The time should roughly double when $n$ doubles (which it does).

```{r}
sapply(c(1e5, 2e5, 4e5, 8e5), function(n) {
    d <- sim_cox(n, beta = 1)
    system.time(for (k in 1:10) pl(1, d$x, d$status))[["elapsed"]]
})
```

### 11.7.4

*Compute and plot an estimate of the baseline cumulative hazard function.*

We use the Breslow estimator from Section 11.4.2,
$$
\hat{\Lambda}_0(t) = \sum_{i: T_i \leq t} \frac{\delta_i}{\sum_{j \geq i} e^{\hat{\beta} X_j}},
$$
which is a cumulative sum of the $\delta_i / \Psi_i^{(0)}(\hat{\beta})$-s, since the data are sorted. We use the data set simulated with $\beta = 1$.

```{r}
beta_hat <- tail(newton_1$beta, 1)
breslow <- data.frame(
    time   = dat1$time,
    cumhaz = cumsum(dat1$status / rev(cumsum(rev(exp(beta_hat * dat1$x)))))
)

ggplot(breslow, aes(time, cumhaz)) +
    geom_step() +
    geom_abline(slope = 1, color = "blue", linetype = "dashed") +
    xlab("Time") +
    ylab("Baseline cumulative hazard")
```

The baseline corresponds to $X = 0$, and the dashed line is the true baseline cumulative hazard $\Lambda_0(t) = t$. The estimate follows it closely for small $t$ and deviates more for large $t$, where few individuals are at risk.

### 11.7.5

*Compute $\hat{\beta}$ using the `coxph()` function in R. Compute also a standard 95% confidence interval for $\beta$. Compare this interval with a (partial) likelihood interval based on $\ell$.*

```{r}
cox_fit <- coxph(Surv(time, status) ~ x, data = dat1)
c(newton = beta_hat, coxph = unname(coef(cox_fit)))
c(newton = pl(beta_hat, dat1$x, dat1$status)$l, coxph = cox_fit$loglik[2])
```

`coxph()` gives the same estimate and the same maximal partial log-likelihood as our implementation. Its baseline cumulative hazard (at $X = 0$) is also the same as our Breslow estimate:

```{r}
max(abs(basehaz(cox_fit, centered = FALSE)$hazard - breslow$cumhaz))
```

The standard 95% confidence interval is $\hat{\beta} \pm 1.96 \cdot \widehat{\mathrm{se}}$ with $\widehat{\mathrm{se}} = (-\ell''(\hat{\beta}))^{-1/2}$.

```{r}
confint(cox_fit)
```

The partial likelihood interval consists of the $\beta$-s with
$$
2 \big(\ell(\hat{\beta}) - \ell(\beta)\big) \leq q_{0.95}(\chi^2_1),
$$
and we find its endpoints with `uniroot()`.

```{r}
l_hat <- pl(beta_hat, dat1$x, dat1$status)$l
se    <- sqrt(-1 / pl(beta_hat, dat1$x, dat1$status)$d2l)  # same as coxph's std. error
comb  <- function(beta) {
    2 * (l_hat - pl(beta, dat1$x, dat1$status)$l) - qchisq(0.95, df = 1)
}
c(
    lower = uniroot(comb, c(beta_hat - 5 * se, beta_hat), tol = 1e-10)$root,
    upper = uniroot(comb, c(beta_hat, beta_hat + 5 * se), tol = 1e-10)$root
)
```

```{r}
beta_grid <- seq(beta_hat - 4 * se, beta_hat + 4 * se, length.out = 200)
data.frame(
    beta      = beta_grid,
    partial   = sapply(beta_grid, function(b) 2 * (l_hat - pl(b, dat1$x, dat1$status)$l)),
    quadratic = (beta_grid - beta_hat)^2 / se^2
) |>
    tidyr::pivot_longer(-beta, names_to = "type") |>
    ggplot(aes(beta, value, color = type)) +
    geom_line() +
    geom_hline(yintercept = qchisq(0.95, df = 1), linetype = "dashed") +
    xlab(expression(beta)) +
    ylab("")
```

The two intervals are almost identical, and both contain the true value $\beta = 1$. The partial likelihood interval is only slightly shifted to the right, because $\ell$ is very close to quadratic around $\hat{\beta}$.
