---
title: "Week 6 Monday"
author: "Clara Rosengaard Groth"
date: "2026-10-02"
format:
  html:
    html-math-method: mathjax
---

## Exercise 11.2

Show that without censoring, the Kaplan–Meier estimator equals the empirical survival function

$$\hat{S}(t)=\frac{1}{n}\sum_{i=1}^n \mathbf{1}(T_i^* > t).$$

When we do not censor, this is essentially taking $C_i \to \infty,\quad \forall i\in\{1,\ldots,n\},$ as we observe every subject until the event occurs. Then
$$
T_i = \min\{T_i^*,C_i\}=T_i^*, \quad \text{and}
$$
$$
\delta_i=\mathbf{1}(T_i^*\leq C_i)=1,
$$
for $i=1,\ldots,n$.

By definition 11.1, the Kaplan–Meier estimator is defined as

$$
\hat S(t) = \prod_{s\leq t}\left( 1-\frac{\Delta N(s)}{Y(s)} \right),
$$

where $\Delta N(s)$ is the number of deaths at time $s$. We allow for ties and define

$$s_1 < s_2 < \cdots < s_k \leq t < s_{k+1} < \cdots < s_m$$

as distinct (not individual) survival times. This means that $m\leq n$, because we allow for multiple deaths/events to occur at the same time. Note that

$$
Y(s_1) = \sum_{i=1}^n \mathbf{1}(s_1\leq T_i) = n,
$$

$$
Y(s_2) = \sum_{i=1}^n \mathbf{1}(s_2\leq T_i) = n-\Delta N(s_1) = Y(s_1)-\Delta N(s_1),
$$

and, realize that in general,

$$
Y(s_k)=\sum_{i=1}^n \mathbf{1}(s_k\leq T_i) = n-\sum_{j=1}^{k-1}\Delta N(s_j) = 
Y(s_{k-1})-\Delta N(s_{k-1}).
$$

Therefore,

$$
\Delta N(s_k)=Y(s_k)-Y(s_{k+1}),
$$

such that the factors of the Kaplan-Meier can be rewritten as

$$
1-\frac{\Delta N(s_k)}{Y(s_k)}=1-\frac{Y(s_k)-Y(s_{k+1})}{Y(s_k)}=\frac{Y(s_{k+1})}{Y(s_k)}.
$$

Then, by expanding the product and seeing that it telescopes, 

$$
\begin{aligned}
\hat S(t)
&=
\prod_{s_k\leq t}
\left(
\frac{Y(s_{k+1})}{Y(s_k)}
\right)\\
&=
\frac{Y(s_2)}{Y(s_1)}
\frac{Y(s_3)}{Y(s_2)}
\cdots
\frac{Y(s_{k+1})}{Y(s_k)}\\
&=
\frac{1}{Y(s_1)}Y(s_{k+1})\\
&=
\frac{1}{n}
\sum_{i=1}^n
\mathbf{1}(s_{k+1}\leq T_i)\\
&=
\frac{1}{n}
\sum_{i=1}^n
\mathbf{1}(t<T_i).
\end{aligned}
$$

The last equality follows from the fact that $t<s_{k+1}  \Longrightarrow \{t<T_i\}=\{s_{k+1}\leq T_i\},$ since $s_{k+1}$ is the next distinct event time after $t$.

Reading trough the equalities we have shown that,

$$
\hat S(t) = \frac{1}{n}\sum_{i=1}^n \mathbf{1}(T_i^*>t),
$$

since there is no censoring and $T_i=T_i^*$.


## Exercise 11.3

```{r, include=FALSE}
#loading libraries
library(RwR)
library(ggplot2)
library(gridExtra)
library(survival)
library(broom)
library(splines)
library(dplyr)
theme_set(theme_bw())
```


The variable `hx` indicates if the patient has a history of cardiovascular disease. You should ignore the treatment variable `rx`. 

```{r}
data("prostate", package = "RwR")
prostate %>% head()
```


### 11.3.1
Fit a proportional hazards model using the two variables `age` and `hx` as predictors.

We fit a Cox's proportional hazards model in accordance with example 11.10 using the R-function `coxph()`. 

```{r}
prostate_cox <- coxph( 
    Surv(dtime + 0.5, status != "alive") ~ age + hx, 
    data = prostate, 
)
prostate_cox %>% tidy()

```

Note that use the exact same data as in the example, and we add $0.5$ to all times, to avoid numerical issues. It is explained in Chapter 10, that some of the patients have survived less than one month, some survival times are registered as $0$. 


### 11.3.2

Make one or more residual plots for the model you have fitted. Does it fit the data?

Section 11.5 discusses different types of residuals for survival regression models, where primarily the Cox-Snell and the Martingales residuals are covered. We therefore plot both types of residuals, and use the code from section 11.5.4 directly.

```{r}
p1 <- augment(prostate_cox, data = prostate, type.predict = "expected") |>
    survfit(Surv(.fitted, status != "alive") ~ 1, data = _) |>
    summary(data.frame = TRUE) |>
    ggplot(aes(time, cumhaz)) + 
    geom_abline(slope = 1, color = "blue") +
    geom_point() +
    xlab("Cox-Snell residuals") + 
    ylab("Cumulative hazard")


p2 <- augment(prostate_cox) |> 
    ggplot(aes(.fitted, .resid)) +
    geom_point(alpha = 0.4) + 
    geom_smooth() + 
    xlab("Fitted values") +
    ylab("Martingale residuals")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

Based on lemma 11.3 we expect the Cox-Snell residuals $\hat{\Lambda}_i(T_i)$  to approximately follow an exponential distribution (for no censoring). Therefore, we compare the estimated Cox-Snell residuals to the cumulative hazard for the exponential distribution, which is $\Lambda(t)=t$. The residuals should lie approximately on the straight line with slope $1$. 

The use of the martingale residuals is justified by lemma 11.4, which implies that the martingale residuals have mean $0$ if $\hat{\Lambda}_i$ is the true cummulative hazard. 

From the residual plots we see that the model is somewhat misspecified. The Cox-Snell residuals does not fall on the straight line for larger residuals. Also, the martingale residuals does not seem to have mean $0$, and the smoothed line is a bit curved.


### 11.3.3 

*Test the hypothesis that survival is not associated with* `hx` *given* `age`.

If we want to test the hypothesis that survival is not associated with `hx` given `age`, the null hypothesis is $$H_0 : \beta_{\operatorname{hx}} = 0. $$ To test this we need to compare the full model with both `hx` and `age`, against a smaller model only containing `age`. The two models are clearly nested, and we can use `anova()`.  

```{r}
prostate_cox_small <- coxph( 
    Surv(dtime + 0.5, status != "alive") ~ age, 
    data = prostate, 
)
```

```{r}
anova(prostate_cox_small, prostate_cox, test = "LRT")
```

We can reject the null hypothesis and conclude that based on a likelihood ratio test survival is indeed associated with a history of cardiovascular disease given age.

## Exercise 11.4

Let $T_1^*, \ldots, T_n^*$ be independent survival times for $n$ patients and let $X_i \in \{0,1\}$ indicate if the $i$-th patient is treated $X_i = 1$ or untreated $X_i = 0$. We are interested in estimating the effect

$$
\rho =
\frac{\operatorname{median}(T_i^* \mid X_i = 1)}
     {\operatorname{median}(T_i^* \mid X_i = 0)}.
$$

Assume

$T_i^* \sim \operatorname{Weibull}(\alpha_{X_i}, \gamma)$ with hazard rate $\lambda_i(t)=\alpha_{X_i}\gamma t^{\gamma-1},$ where $\gamma$ is the shape parameter and
$\alpha_{X_i}$ is the scale parameter.

### 11.4.1

Show that 
$$
\rho = \left(\frac{\alpha_0}{\alpha_1}\right)^{1/\gamma}.
$$

From the definition of a median,

$$P(T_i^* \leq m_{X_i} \mid X_i = x) = \frac{1}{2}.$$

$T^*$ has distribution function $F$ and survival function $S$, for which $S(t) := 1 - F(t).$
This implies 
$$S(m_{X_i}) = \frac{1}{2}.$$

From (11.1) and (11.2), we have $S(t)=\exp(-\Lambda(t))=\exp(-\alpha t^\gamma),$ where the last equality follows from Example 11.3 for a Weibull distribution. Then

$$S(m_{X_i})= \exp\left(-\alpha_{X_i}m_{X_i}^{\gamma}\right) = \frac{1}{2}.$$

Then

$$
m_{X_i}= \left(\alpha_{X_i}^{-1}\right)^{1/\gamma} \left(\log(2)\right)^{1/\gamma}.
$$

Which implies,

$$
\rho
=
\frac{
\left(\alpha_1^{-1/\gamma}\log(2)\right)^{1/\gamma}
}{
\left(\alpha_0^{-1/\gamma}\log(2)\right)^{1/\gamma}
}
=
\left(\frac{\alpha_0}{\alpha_1}\right)^{1/\gamma}.
$$

Note that this exercise can also be solved by looking up the median for a Weibull distribution on e.g. Wikipedia, but beware of the parametrization. 

### 11.4.2

Show that the Weibull model can be formulated as a proportional hazards model in terms of $\eta_i = \beta_0 + \beta_1 X_i,$ and relate $\beta_0$ and $\beta_1$ to $\alpha_0$ and $\alpha_1$.

The proportional hazards model is defined by (definition 11.5)
$$\lambda(t \mid X_i)=\lambda_0(t)e^{\eta_i},$$

which gives us

$\lambda(t \mid X_i = 1)=\lambda_0(t)e^{\beta_0+\beta_1},$ and $\lambda(t \mid X_i = 0)=\lambda_0(t)e^{\beta_0}.$

Therefore,

$$
\frac{\lambda(t \mid X_i = 1)}{\lambda(t \mid X_i = 0)} = \frac{e^{\beta_0+\beta_1}}{e^{\beta_0}}.
$$

This is a consequence of the model assumption for proportional hazards models that the effect of $X$ enters multiplicatively only through $e^\eta$. For the Weibull model,

$$
\frac{\lambda(t \mid X_i = 1)}{\lambda(t \mid X_i = 0)} = \frac{\alpha_1 \gamma t^{\gamma-1}}{\alpha_0 \gamma t^{\gamma-1}} = \frac{\alpha_1}{\alpha_0}.
$$

Therefore,

$$
\frac{\alpha_1}{\alpha_0}
=
\frac{e^{\beta_0+\beta_1}}{e^{\beta_0}},
$$

with $\alpha_0 = e^{\beta_0},\qquad\alpha_1 = e^{\beta_0+\beta_1}.$

Finally,
$$
\rho=\left(\frac{e^{\beta_0}}{e^{\beta_0+\beta_1}}\right)^{1/\gamma} = e^{-\beta_1/\gamma}.
$$

### 11.4.3

```{r}
data("testWei")
```

Compute and plot the Kaplan-Meier estimates of the survival function within each of the two treatment groups for the simulated `testWei` data set.

First we check if there are any times very close to $0$. If there is, we might need to add a small constant to the time, as it is done throughout Chapter 11, to avoid numerical issues. 
```{r}
summary(testWei$time)
```

We continue the analysis without shifting the time index. 

The Kaplan-Meier estimator of the survival function can be fitted using `survfit()` as explained in Section 11.2.3. Note that we use `Surv()` function to include censoring status. In this dataset censoring status is encoded in `status` where $1$ corresponds to time to event observed and $0$ corresponds to censored. Note also that we include the treatment variable `x` in the formula, to estimate the survival function for each of the treatment groups.

```{r}
Weibull_KM <- survfit(
    Surv(time, status == 1) ~ x, 
    data = testWei,
)
```

We plot the estimated survival functions as in Section 11.2.3.

```{r}
p1 <- summary(Weibull_KM, data.frame = TRUE) |> 
        ggplot(aes(time, surv, color = strata)) + 
        geom_step() +
        ylim(0, 1) +
        xlab("Time") + 
        ylab("Survival probability")+
        theme(legend.position = "top")

p1
```



### 11.4.4

Use the Weibull model with fixed $\gamma = 2$ to estimate $\beta_0$ and $\beta_1$. Compare the resulting estimated survival functions from the Weibull model with the Kaplan-Meier estimates.

From the previous subexercises we know that the proportial hazards Weibull model satisfies
$$\lambda_i(t ) = \gamma t^{\gamma -1} e^{\eta_i}, \quad \eta_i = \beta_0 + \beta_1 X_i.$$

Example 11.8 shows that, when $\gamma$ is fixed, the survival log-likelihood for the Weibull model is up to a constant the log-likelihood for the Poisson model with log-link for the $\delta_i$-s. This means that 
$$\delta_i \sim \operatorname{Poisson(\mu_i)}, \quad \quad \mu_i = T_i^{\gamma}e^{\eta_i}.$$

Taking the logarithm we get 

$$\log(\mu_i) = \gamma \log(T_i) + \eta_i.$$

This shows how implementations of generalized linear models (with Poisson regression models), can be used to fit the Weibull proportional hazards model (for fixed $\gamma$) where the survival times enters as an offset term. We can therefore estimate $\beta_0$ and $\beta_1$ using `glm()`, fitting exactly the model given above:

```{r}
gamma <- 2
PoissonGLM <- glm((status == 1) ~ offset(log(time)*gamma) + x, 
    data = testWei, 
    family = "poisson"
)
beta_hat <- PoissonGLM %>% coefficients() 
beta_hat
```

We are then also asked to compare the resulting estimated survival functions from the Weibull model with the Kaplan-Meier estimates. From subexercise 11.4.1 we know that $\alpha_0 = e^{\beta_0}$ and $\alpha_1 = e^{\beta_0+\beta_1},$ which results in the following survival functions
$$S(t \mid X_i = 0) = \exp(-e^{\beta_0}t^{\gamma}),$$

$$S(t \mid X_i = 1) = \exp(-e^{\beta_0 + \beta_1}t^{\gamma}).$$

This can be straightforwardly implemented in R, and plotted using the KM-figure from subexercise 11.4.3.

```{r}
time <- seq(0, 3.5, 0.01)

S_hat_x0 <- exp(-exp(beta_hat[1]) * time^gamma)
S_hat_x1 <- exp(-exp(beta_hat[1] + beta_hat[2]) * time^gamma)

Weibull_curves <- data.frame(
  time = time,
  S_hat_x0 = S_hat_x0,
  S_hat_x1 = S_hat_x1
)

p1 + 
  geom_line(data = Weibull_curves, aes(x = time, y = S_hat_x0),
            linetype = "dashed", colour = "darkred", inherit.aes = FALSE) +
  geom_line(data = Weibull_curves, aes(x = time, y = S_hat_x1), colour = "blue3",
            linetype = "dashed", inherit.aes = FALSE)
```

From the figure above it seems like the estimated survival functions from the Weibull model with $\gamma=2$ corresponds well to the Kaplan-Meier estimates. There are some deviances, which we would expect due to sampling variability and the fact that KM is a nonparametric estimator.


### 11.4.5

Compute the MLE $\hat{\rho}_{\operatorname{MLE}}$ of $\rho$ based on the Weibull model (with fixed $\gamma =2$) and compute a 95% confidence interval for $\rho$. Compute also, based on the Kaplan-Meier estimates, a direct estimate, $\hat{\rho}_{\operatorname{KM}}$ of $\rho$.

Using the MLE estimates of $\beta_0$ and $\beta_1$ from the previous subexercise 11.4.4, we can use the formula from exercise 11.4.2 to obtain
$$\hat{\rho}_{\operatorname{MLE}} = e^{-\frac{\hat{\beta}_1}{\gamma}}$$

```{r}
beta1_hat_MLE <- beta_hat[2]
rho_hat_MLE <- exp((-beta1_hat_MLE/2))
rho_hat_MLE
```

And a confidence interval for $\hat{\beta}_1$ can be computed using the estimated standard error for $\beta_1$ in our `glm()` fit, and then transformed by $f(\beta)=e^{-\beta/2}$.

```{r}
beta1_se <- summary(PoissonGLM)$coefficients[2,2]
beta1_CI <- beta1_hat_MLE + c(1,-1)*1.96*beta1_se
rho_CI <- exp(-beta1_CI/2)
rho_CI
```

We also need to compute the $\hat{\rho}_{\operatorname{KM}}$ based on the Kaplan Meier estimates. Note that the medians can be extracted directly from the `survfit()` object. 

```{r}
Weibull_KM
```


```{r}
rho_hat_KM <- summary(Weibull_KM)$table[, "median"][2]/summary(Weibull_KM)$table[, "median"][1]
rho_hat_KM
```

From the exercise description we know that the true value of $\rho$ is approximately $1.649$. Note that $\hat{\rho}_{\operatorname{MLE}}$ is much closer to $\rho \approx 1.649$ than $\hat{\rho}_{\operatorname{KM}}$. This difference may be due to sampling variability, which we will examine in the next subexercise.


### 11.4.6

Write an R program that simulates a data set (including the censoring) using the same setup as for the `testWei` data set described above. Simulate $B = 200$ data sets and compute for each data set the estimates $\hat{\rho}_{\operatorname{KM}}$ and $\hat{\rho}_{\operatorname{MLE}}$ as above and compare their distributions.

First we simulate the data set. We know that time to event $T_i^*$ is Weibull distributed with shape $\gamma = 2$ and scale $\alpha_0 = e^{\beta_0}$ for $X_i=0$ and $\alpha_1 = e^{\beta_0 + \beta_1}$ for $X_i=1$, but be aware of the parametrization used in `rweibull()`, and when in doubt please consult the documentation. R uses the same parametrization as the "standard" parametrization from Wikipedia (https://en.wikipedia.org/wiki/Weibull_distribution), where in this exercise we use the "first option" parametrization.

The dataset is simulated in accordance with the information given in the exercise description; $X_1 = .. = X_{100} = 0$ and $X_{101} = ... X_{200} = 1$ and the censoring are uniformly distributed on $[0,4]$.

```{r}
set.seed(42)
N <- 200 # number of patients
B <- 200 # number of repetitions
beta0 <- 0
beta1 <- -1
gamma <- 2
datasets <- vector("list", B)
rho_hat_KM <- rep(NA, B)
rho_hat_MLE <- rep(NA, B)

for (i in 1:B){
  X <- c(rep(0, 100), rep(1, 100))
  alpha <- ifelse(X == 0, exp(beta0), exp(beta0 + beta1))
  T_star <- rweibull(N, shape = gamma, scale = alpha^(-1/gamma))
  C <- runif(N, min = 0, max = 4)
  T <- pmin(T_star, C)
  status <- ifelse(T_star <= C, 1, 0) # censoring
  
  datasets[[i]] <- data.frame(
    x = X,
    time = T,
    status = status
  )
  
  # Kaplan Meier estimator
  Weibull_KM_i <- survfit(
    Surv(time, status == 1) ~ x, 
    data = datasets[[i]])
  rho_hat_KM[i] <- summary(Weibull_KM_i)$table[, "median"][2]/summary(Weibull_KM_i)$table[, "median"][1]
  
  # MLE estimator
  GLM_coefs_i <- glm((status == 1) ~ offset(log(time)*gamma) + x, 
    data = datasets[[i]], family = "poisson") %>% coefficients() 
  rho_hat_MLE[i] <- exp((-GLM_coefs_i[2]/gamma))
} 

```


And we plot the simulated values using a violinplot. 

```{r}

estplot_data <- data.frame(
  estimator = c(rep("KM", length(rho_hat_KM)), rep("MLE", length(rho_hat_MLE))),
  rho_hat = c(rho_hat_KM, rho_hat_MLE))

ggplot(estplot_data, aes(x = estimator, y = rho_hat, fill = estimator)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.8) +
  geom_hline(yintercept = exp(-beta1 / gamma), linetype = "dashed") +
  labs(
    title = "Rho estimates",
    x = "Estimator",
    y = expression(hat(rho))
  ) +
  theme_minimal()
```

We can also summarize the simulated distributions:

```{r}
data.frame(
  estimator = c("KM", "MLE"),
  mean = c(mean(rho_hat_KM, na.rm = TRUE), mean(rho_hat_MLE, na.rm = TRUE)),
  sd = c(sd(rho_hat_KM, na.rm = TRUE), sd(rho_hat_MLE, na.rm = TRUE))
) 
```


The dashed line on the plot is the true value of $\rho$. We see that the the distributions for the two estimators are fairly alike. They both have a mean close to the true value, however there is a larger variance for the Kaplan-Meier estimator. This is perhaps not that surprising; the MLE assumes the correctly specified Weibull model, whereas KM is nonparametric. By construction in this simulation, the Weibull assumption is correct, and the parametric approach is therefore more efficient.
