---
title: "Week 1: Wednesday Solutions"
---

```{r, include = FALSE}
library(RwR)
library(ggplot2)
library(tibble)
library(broom)
library(splines)
library(gridExtra)

theme_set(theme_bw()) # plotting theme is always theme_bw()
```

## Exercise 1 - 2.2
This exercise is based on a dataset relating the mean verbal SAT scores for each of the 50 US states to the percentage of students taking the test. The data comes with the RwR package.

```{r}
library(RwR)
data(sat, package = "RwR")
```

We take a brief look at the dataset

```{r}
summary(sat)
p0 <- ggplot(data = sat, aes(x = percent, y = verbalScore)) + 
    geom_point()
p0
```

#### 1. Fit a linear regression model (with an intercept) of the state SAT score as a function of percentage of students. Add the fitted line to the plot.

We start by fitting the model

```{r}
fit <- lm(verbalScore ~ percent, data = sat)
summary(fit)
```

We visiualize the fitted linear regression model

```{r}
p0 + geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
```

Also an alternative solutions using `predict`, i.e. predicting values from the model
```{r}
fit.predict <- tibble(fit = predict(fit, newdata = sat))

p0 + geom_line(aes(y = fit.predict$fit), color = "blue") 
```

#### 2. Fit a piecewise affine function with the “bend” at 50% of the students taking the test. How can you formally test the linearity assumption in Question 1 within this larger model?

Recall the hint: You want to fit a linear model with a model matrix 
with three columns representing an intercept and two piecewise linear functions. You can, for instance, generate those piecewise linear functions by an ``ifelse()`` construction in the formula. Try different choices of columns. Is there one choice particularly suitable for testing the linearity assumption?

We need to fit a piecewise affine mean function with a bend at $50\%$ students taking the test. That is, we have to fit a mean function with two slopes: one for values below $50\%$ and for for after $50\%$. 

We construct the model matrix with three columns: 1) the intercept; 2) the column $x$; and 3) the column $\operatorname{max}(x - 50, 0)$.

So if $x \leq 50$ then the third column is $0$ and there is only contribution from the intercept and the second column and if $x > 50$ then there is another contribution to the slope(corresponding to a change in the slope). 

It corresponds to the mean function
\begin{align*}
m(x) &= \mathbf E(Y \mid X = x) \\
     &= \beta_0 + \beta_1 x + \beta_2 (x - 50)_+ \\
     &= \begin{cases}
\beta_0+\beta_1x, & x\leq 50,\\
\beta_0+\beta_1x+\beta_2(x-50), & x>50 \end{cases} \\
&= \begin{cases} \beta_0 + \beta_1 x & x \leq 50 \\ (\beta_0 - 50\beta_2) + (\beta_1 + \beta_2) x & x > 50 \end{cases}.
\end{align*}

Note how the slope changes if $x > 50$ compared to if $x \leq 50$, or in other words, $\beta_2$ changes the slope. $m(x)$ is also continuous in $x$. 

We define the two needed linear functions

```{r}
f1 <- function(x) x
f2 <- function(x) ifelse(x <= 50, 0, x - 50) # equivalently pmax(x - 50, 0)
```

We visualize the 'bend' implied by the the functions
```{r}
bend_data <- data.frame(
    percent = rep(sat$percent, 2), 
    y = c(f1(sat$percent), f2(sat$percent)), 
    variable = rep(c("f1", "f2"), each = nrow(sat))
)

ggplot(bend_data, aes(x = percent, y = y, color = variable)) +
  geom_line() 
```

Now, we fit the model
```{r}
piecewise_fit <- lm(verbalScore ~ f1(percent) + f2(percent), data = sat)
summary(piecewise_fit)
```

yielding the (approximate) estimated mean function
$$
\widehat m(x)=
\begin{cases}
497.785-1.635x, & x\leq 50,\\
383.463+0.652x, & x>50.
\end{cases}
$$

We see that the fitted slopes change from $-1.635$ below 50 to $0.652$ above 50.

Setting $\beta_2 = 0$ removes the 'bend' term and gives $m(x) = \beta_0 + \beta_1 x$ for every $x$, the model from Question 1. 

We can also see this using the model matrices. The model matrix from Question 1 has columns 1) intercept; and 2) $x$. Since we add another column to the model matrix in this question, the model from Question 1 is nested in the current one (the column spaces are nested). 

Formally we test the hypothesis
$$
H_0: \beta_2 = 0
$$
against the alternative that $\beta_2$ is non-zero. We can test this directly with a $t$-test, by looking at the summary printed from the model above. We find $t = 5.806$ and $p=5.28\times10^{-7}$. Alternatively, we can use an $F$-test
```{r}
anova(fit, piecewise_fit)
```

The test gives $F=33.71$ and $p=5.28\times10^{-7}$. Note that because the $H_0$ is just a one-dimensional parameter restriction, we have $F = t^2$ in this case, and that the $p$-values are the same. We reject the hypothesis $H_0$, that is we reject that $\beta_2 = 0$, or in words: The data provides evidence for a change of slopes at 50. 


## Exercise 2 - 2.5
This exercise relates the empirical correlations between the predictor variables to the variance of the least squares estimator. The setup is as follows. The matrix $\mathbf{X}$ denotes an $n \times p$ matrix of predictor variables, $\mathbf{1}$ denotes an  $n$-dimensional column vector of ones and $\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}$ denotes the $n \times (p + 1)$ matrix with the first column being $\mathbf{1}$ and the remaining $n \times p$ block being $\mathbf{X}$. We let
$$
\bar X = \frac{1}{n} \mathbf{X}^\top \mathbf{1}
$$
denote the $p$-dimensional vector of mean values for the predictor (the empirical average of each predictor), and we let
 $$
 \hat \Sigma = \frac{1}{n}(\mathbf{X} - \mathbf{1} \bar X^\top)^\top (\mathbf{X} - \mathbf{1} \bar X^\top)
 $$
denote the empirical covariance matrix. It is assumed that $\hat \Sigma$ is invertible.

#### 1. Show that
$$
n \hat \Sigma = \mathbf{X}^\top \mathbf{X} - n\bar X \bar X^\top.
$$

We simply plug into the definitions and along the way we use that $\mathbf{1}^\top \mathbf{1} = n$. We have
\begin{align*}
n\hat\Sigma
&= (\mathbf X-\mathbf 1\bar X^\top)^\top
   (\mathbf X-\mathbf 1\bar X^\top) \\
&= (\mathbf X^\top-\bar X\mathbf 1^\top)
   (\mathbf X-\mathbf 1\bar X^\top) \\
&= \mathbf X^\top\mathbf X
   -\mathbf X^\top\mathbf 1\bar X^\top
   -\bar X\mathbf 1^\top\mathbf X
   +\bar X\mathbf 1^\top\mathbf 1\bar X^\top \\
&= \mathbf X^\top\mathbf X
   -n\bar X\bar X^\top
   -n\bar X\bar X^\top
   +n\bar X\bar X^\top \\
&= \mathbf X^\top\mathbf X-n\bar X\bar X^\top.
\end{align*}

#### 2. Use block matrix inversion to show that
$$
\begin{pmatrix} \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}^\top \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix} \end{pmatrix}^{-1} = \begin{pmatrix} \ast & \ast \\ \ast & \frac{1}{n}\hat \Sigma^{-1} \end{pmatrix}.
$$

First we recall the block matrix inversion formula
$$
P = \begin{pmatrix} A & B \\ C & D \end{pmatrix}^{-1} = \begin{pmatrix} \ast & \ast \\ \ast & (D - C A^{-1} B)^{-1} \end{pmatrix},
$$
which holds whenever $A$ is invertible, and the Schur complement of $A$ in $P$ is invertible.

Now let $\mathbf{Z} = \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}$. Then
\begin{align*}
\mathbf{Z}^\top \mathbf{Z} &= \begin{pmatrix} \mathbf{1}^\top \\ \mathbf{X}^\top \end{pmatrix} \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix} \\
&= \begin{pmatrix} \mathbf{1}^\top \mathbf{1} & \mathbf{1}^\top \mathbf{X} \\ \mathbf{X}^\top \mathbf{1} & \mathbf{X}^\top \mathbf{X} \end{pmatrix} \\
&= \begin{pmatrix} n & \mathbf{1}^\top \mathbf{X} \\ \mathbf{X}^\top \mathbf{1} & \mathbf{X}^\top \mathbf{X} \end{pmatrix}.
\end{align*}


Thus, using block inversion, recalling the definition of $\bar X$, and the result from the previous exercise, we find that
\begin{align*}
\begin{pmatrix} \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}^\top \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix} \end{pmatrix}^{-1} &= (\mathbf{Z}^\top \mathbf{Z})^{-1} \\
&= (\mathbf{X}^\top \mathbf{X} - \mathbf{X}^\top \mathbf{1} \frac{1}{n} \mathbf{1}^\top \mathbf{X})^{-1} \\
&= (\mathbf{X}^\top \mathbf{X} - n \bar X \frac{1}{n} \bar X^\top)^{-1} \\
&= (\mathbf{X}^\top \mathbf{X} - n \bar X \bar X^\top)^{-1} \\
&=\frac{1}{n} \hat\Sigma^{-1}.
\end{align*}

We have to assume that we have observed some data, that is $n>0$ so that $n$ is invertible and we have also assumed that $\hat \Sigma$ is invertible, so the block inversion result can be applied and the above calculation stands true. 

#### 3. Show that if we fit a linear model with an intercept (which is denoted $\beta_0$) using the least squares estimator, then the variance of $\hat\beta_i$, conditionally on $\mathbf{X}$, for $i = 1, \dots, p$ is $\frac{\sigma^2}{n}(\hat\Sigma^{-1})_{ii}$

Write $\beta = (\beta_0, \beta_1, \dots, \beta_p)$ such that the first coordinate corresponds to the intercept. The belonging model matrix is $\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}$. 

Under the assumption of \textbf{A1}, \textbf{A2}, \textbf{A4}, and that $\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}$ has full rank, Theorem 2.2 provides the variance of the ordinary least squares estimator
$$
\operatorname{Var}(\hat\beta \mid \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}) = \sigma^2 (\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}^\top \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix})^{-1}. 
$$

Using the result from the previous exercise, 
$$
\operatorname{Var}(\hat\beta \mid \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}) = \sigma^2 \begin{pmatrix} \ast & \ast \\ \ast & \frac{1}{n}\hat \Sigma^{-1} \end{pmatrix}.
$$

Now it is just a matter of picking out the correct components of the covariance matrix, and we get, for each $i = 1, \dots, p$, that
$$
\operatorname{Var}(\hat\beta_i \mid \begin{pmatrix} \mathbf{1} & \mathbf{X}\end{pmatrix}) = \frac{\sigma^2}{n} (\hat \Sigma^{-1})_{ii}.
$$

Note that it does not matter whether or not we condition on $\mathbf{X}$ or $\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}$ since $\sigma(\mathbf{X}) = \sigma(\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix})$. This proves the result. 

Let $\hat\Sigma_{-1}$ be the empirical covariance matrix with the first row and column removed. Furthermore, write $\hat\Sigma$ in block form as
$$
\hat \Sigma = \begin{pmatrix} \hat\sigma_1^2 & \hat\gamma^\top \\
\hat\gamma & \hat\Sigma_{-1}
\end{pmatrix}.
$$

#### 4. Use block inversion (again) to show that
$$
(\hat\Sigma_{11})^{-1} = \frac{1}{\hat\sigma_1^2 - \hat\gamma^\top (\hat\Sigma_{-1})^{-1} \hat\gamma}. 
$$

We have to assume that $\hat\sigma_1 \neq 0$ to apply the block inversion formula. Because 


Furtermore, we also need $\hat\sigma_1^2 - \hat\gamma^\top (\hat\Sigma_{-1})^{-1} \hat\gamma$ to be invertible, which is equivalent to $\hat\sigma_1^2 \neq \hat\gamma^\top (\hat\Sigma_{-1})^{-1} \hat\gamma$ -- can this be shown or do we have to assume that?. 

Assuming that, block inversion applied to the block-decomposition of $\Sigma$ given above yields the result.

#### 5. Show that in the case $p = 2$, the formula above reduces to
$$
(\hat \Sigma^{-1})_{11} = \frac{1}{\hat\sigma_1^2 (1 - \widehat{\operatorname{corr}}(X_1, X_2)^2)}.
$$

Recall the definition of the correlation between two random variables (with finite variance) $X$ and $Y$
$$
\operatorname{corr}(X, Y) = \frac{\operatorname{cov}(X, Y)}{\sqrt{\operatorname{Var}(X) \operatorname{Var}(Y)}}
$$
whereby
$$
\operatorname{corr}(X, Y)^2 = \frac{\operatorname{cov}(X, Y)^2}{\operatorname{Var}(X) \operatorname{Var}(Y)}.
$$

If $p = 2$ then there are two (univariate) predictors $X_1$ and $X_2$, and $\hat\Sigma_{-1} = \operatorname{Var}(X_2)$ and $\hat \gamma = \widehat{\operatorname{cor}}(X_1, X_2)$. Also, $\operatorname{Var}{X_1} = \sigma_1^2$ and $\operatorname{Var}{X_2} = \sigma_2^2$ and they have empirical variances denoted by a $\hat\cdot$, and we wil use the $\operatorname{Var}$ and $\sigma^2$ interchangeably. 

The formula from the previous exercise new yields
\begin{align*}
(\hat\Sigma_{-1})_{11} &= \frac{1}{\hat\sigma_1^2 - \widehat{\operatorname{cov}}(X_1, X_2) \frac{1}{\widehat{\operatorname{Var}}(X_2)}\widehat{\operatorname{cov}}(X_1, X_2)} \\
&= \frac{1}{\hat\sigma_1^2 - \frac{\widehat{\operatorname{cov}}(X_1, X_2)}{\widehat{{\operatorname{Var}}(X_2)}}} \\
&= \frac{1}{\hat\sigma_1^2 (1 - \frac{\widehat{\operatorname{cov}}(X_1, X_2)}{\widehat{\operatorname{Var}}(X_2)} \frac{1}{\hat\sigma_1^2})} \\
&= \frac{1}{\hat\sigma_1^2 ( 1 - \widehat{\operatorname{corr}}(X_1, X_2)^2)},
\end{align*}
as desired. 

#### 6. Show that if $\mathbf{X}_1 - \mathbf{1} \bar X_1$ is perpendicular to the remaining columns in $\mathbf{X} - \mathbf{1} \bar X^\top$ , then
$$
(\hat\Sigma^{-1})_{11} = \frac{1}{\hat\sigma_1^2}.
$$

To distinguish columns from scalar entries, write the
$j$th column of the design matrix as $\mathbf X_{\cdot j}$ (as done in the exercise description). There are $p - 1$
remaining columns, indexed by $j = 2, \ldots, p$, so the orthogonality assumption
can be written as

$$
(\mathbf X_{\cdot 1} - \mathbf1 \bar X_1)^\top
(\mathbf X_{\cdot j}-\mathbf1 \bar X_j)=0,
\qquad j= 2, \ldots, p.
$$

The vector $\hat\gamma\in\mathbb R^{p-1}$ contains the
empirical covariances between the first predictor and the remaining predictors.
More precisely, for $j = 2, \ldots, p$,

$$
\hat\gamma_{j-1}
=\frac1n
(\mathbf X_{\cdot1}-\mathbf1\bar X_1)^\top
(\mathbf X_{\cdot j}-\mathbf1\bar X_j).
$$

Thus orthogonality implies $\hat\gamma=\mathbf0_{p-1}$. 

The result from exercise 4 now yields
\begin{align*}
(\hat\Sigma^{-1})_{11} = \frac{1}{\hat\sigma_1^2 - \hat\gamma^\top (\hat\Sigma_{-1})^{-1} \hat\gamma} = \frac{1}{\hat\sigma_1^2}.
\end{align*}


#### 7. Interpret the results obtained in this exercise in terms of how the empirical variance of and empirical correlation between predictor variables affect the variances of the estimators $\hat \beta_i$ for $i = 1, \dots, p$.

For $p = 2$, recall that
$$
\operatorname{Var}(\hat \beta_1 \mid \mathbf{X}) = \frac{\sigma^2}{n} \frac{1}{\hat\sigma_1^2(1 - \widehat{\operatorname{corr}}(X_1, X_2)^2)}.
$$

This is decreasing in if $\hat\sigma_1^2$ increases, and it blows up whenever $\widehat{\operatorname{corr}}(X_1, X_2)$ approaches $1$ or $-1$. Thus, the (conditional) variance of $\hat \beta_1$ is very large if the empirical variance of $X_1$ is very small or if the predictors are very correlated (both positively and negatively). 

For $p > 2$, we see from exercise 6 that the (conditional) variance of $\hat\beta_1$ is minimized when $\mathbf{X}_{\cdot 1} - \mathbf{1} \bar X_1$ is orthogonal to the other centered predictors, because $\hat\Sigma_{-1}$ is positive (semi?)definite and thus $\hat\gamma^\top (\hat\Sigma_{-1})^{-1} \hat\gamma \geq 0$ with equality when the orthogonality assumption holds. 

## Exercise 3 - 3.1

Add the exercise and its solution here.
