---
title: "Week 1: Wednesday Solutions"
author: "Kasper Niss and Asbjørn Risom"
---

```{r, include = FALSE}
library(RwR)
library(ggplot2)
library(tibble)
library(broom)
library(splines)
library(gridExtra)
library(dplyr)

theme_set(theme_bw()) # plotting theme is always theme_bw()
```

## Exercise 1 - 2.2

This exercise is based on a dataset relating the mean verbal SAT scores for each of the 50 US states to the percentage of students taking the test. The data comes with the RwR package.

```{r}
library(RwR)
data(sat, package = "RwR")
```

We take a brief look at the dataset

```{r}
summary(sat)
p0 <- ggplot(data = sat, aes(x = percent, y = verbalScore)) + 
    geom_point()
p0
```

#### 1. Fit a linear regression model (with an intercept) of the state SAT score as a function of percentage of students. Add the fitted line to the plot.

We start by fitting the model

```{r}
fit <- lm(verbalScore ~ percent, data = sat)
summary(fit)
```

We visiualize the fitted linear regression model

```{r}
p0 + geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
```

Also an alternative solution using `predict`, i.e. predicting values from the model
```{r}
fit.predict <- tibble(fit = predict(fit, newdata = sat))

p0 + geom_line(aes(y = fit.predict$fit), color = "blue") 
```

#### 2. Fit a piecewise affine function with the “bend” at 50% of the students taking the test. How can you formally test the linearity assumption in Question 1 within this larger model?

Recall the hint: You want to fit a linear model with a model matrix 
with three columns representing an intercept and two piecewise linear functions. You can, for instance, generate those piecewise linear functions by an ``ifelse()`` construction in the formula. Try different choices of columns. Is there one choice particularly suitable for testing the linearity assumption?

We need to fit a piecewise affine mean function with a bend at $50\%$ students taking the test. That is, we have to fit a mean function with two slopes: one for values below $50\%$ and for for after $50\%$. 

We construct the model matrix with three columns: 1) the intercept; 2) the column $x$; and 3) the column $\operatorname{max}(x - 50, 0)$.

So if $x \leq 50$ then the third column is $0$ and there is only contribution from the intercept and the second column and if $x > 50$ then there is another contribution to the slope (corresponding to a (potential) change in the slope). 

It corresponds to the mean function
\begin{align*}
m(x) &= \mathbf E(Y \mid X = x) \\
     &= \beta_0 + \beta_1 x + \beta_2 (x - 50)_+ \\
     &= \begin{cases}
\beta_0+\beta_1x, & x\leq 50,\\
\beta_0+\beta_1x+\beta_2(x-50), & x>50 \end{cases} \\
&= \begin{cases} \beta_0 + \beta_1 x & x \leq 50 \\ (\beta_0 - 50\beta_2) + (\beta_1 + \beta_2) x & x > 50 \end{cases}.
\end{align*}

Note how the slope changes if $x > 50$ compared to if $x \leq 50$.

We define the two needed linear functions

```{r}
f1 <- function(x) x
f2 <- function(x) ifelse(x <= 50, 0, x - 50) # equivalently pmax(x - 50, 0)
```

We visualize the 'bend' implied by the the functions
```{r}
bend_data <- data.frame(
    percent = rep(sat$percent, 2), 
    y = c(f1(sat$percent), f2(sat$percent)), 
    variable = rep(c("f1", "f2"), each = nrow(sat))
)

ggplot(bend_data, aes(x = percent, y = y, color = variable)) +
  geom_line() 
```

Now, we fit the model
```{r}
piecewise_fit <- lm(verbalScore ~ f1(percent) + f2(percent), data = sat)
summary(piecewise_fit)
```

yielding the (approximate) estimated mean function
$$
\widehat m(x)=
\begin{cases}
497.785-1.635x, & x\leq 50,\\
383.463+0.652x, & x>50.
\end{cases}
$$

We see that the fitted slopes change from $-1.635$ below 50 to $0.652$ above 50.

Fitting the model without the parameter $\beta_2$, removes the 'bend' term and gives $m(x) = \beta_0 + \beta_1 x$ for every $x$, that is the model from Question 1. 

We can also see this using the model matrices. The model matrix from Question 1 has columns 1) intercept; and 2) $x$. Since we add another column to the model matrix in this question, the model from Question 1 is nested in the current one (the column spaces are nested). 

To test the hypothesis of linearity, we can test the hypothesis
$$
H_0: \beta_2 = 0
$$
against the alternative that $\beta_2$ is non-zero. We can test this directly with a $t$-test, by looking at the summary printed from the model above. We find $t = 5.806$ and $p=5.28\times10^{-7}$, and thus data indicate there should be a bend. 

Alternatively, we can use an $F$-test (because the models are nested)
```{r}
anova(fit, piecewise_fit)
```

The test gives $F=33.71$ and $p=5.28\times10^{-7}$. Note that because $H_0$ is just a one-dimensional parameter restriction, we have $F = t^2$ in this case, and that the $p$-values are the same. We reject the hypothesis $H_0$, that is we reject that $\beta_2 = 0$, or in words: The data provides evidence for a change of slopes at 50. 


## Exercise 2 - 2.5

This exercise relates the empirical correlations between the predictor variables to the variance of the least squares estimator. The setup is as follows. The matrix $\mathbf{X}$ denotes an $n \times p$ matrix of predictor variables, $\mathbf{1}$ denotes an  $n$-dimensional column vector of ones and $\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}$ denotes the $n \times (p + 1)$ matrix with the first column being $\mathbf{1}$ and the remaining $n \times p$ block being $\mathbf{X}$. We let
$$
\bar X = \frac{1}{n} \mathbf{X}^\top \mathbf{1}
$$
denote the $p$-dimensional vector of mean values for the predictor (the empirical average of each predictor), and we let
 $$
 \hat \Sigma = \frac{1}{n}(\mathbf{X} - \mathbf{1} \bar X^\top)^\top (\mathbf{X} - \mathbf{1} \bar X^\top)
 $$
denote the empirical covariance matrix. It is assumed that $\hat \Sigma$ is invertible.

#### 1. Show that

$$
n \hat \Sigma = \mathbf{X}^\top \mathbf{X} - n\bar X \bar X^\top.
$$

We simply plug into the definitions and along the way we use that $\mathbf{1}^\top \mathbf{1} = n$. We have
\begin{align*}
n\hat\Sigma
&= (\mathbf X-\mathbf 1\bar X^\top)^\top
   (\mathbf X-\mathbf 1\bar X^\top) \\
&= (\mathbf X^\top-\bar X\mathbf 1^\top)
   (\mathbf X-\mathbf 1\bar X^\top) \\
&= \mathbf X^\top\mathbf X
   -\mathbf X^\top\mathbf 1\bar X^\top
   -\bar X\mathbf 1^\top\mathbf X
   +\bar X\mathbf 1^\top\mathbf 1\bar X^\top \\
&= \mathbf X^\top\mathbf X
   -n\bar X\bar X^\top
   -n\bar X\bar X^\top
   +n\bar X\bar X^\top \\
&= \mathbf X^\top\mathbf X-n\bar X\bar X^\top.
\end{align*}

#### 2. Use block matrix inversion to show that

$$
\begin{pmatrix} \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}^\top \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix} \end{pmatrix}^{-1} = \begin{pmatrix} \ast & \ast \\ \ast & \frac{1}{n}\hat \Sigma^{-1} \end{pmatrix}.
$$

First we recall the block matrix inversion formula
$$
P^{-1} = \begin{pmatrix} A & B \\ C & D \end{pmatrix}^{-1} = \begin{pmatrix} \ast & \ast \\ \ast & (D - C A^{-1} B)^{-1} \end{pmatrix},
$$
which holds whenever $A$ is invertible, and the Schur complement of $A$ in $P$, $D - C A^{-1} B$, is invertible.

We have
\begin{align*}
\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}^\top \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix} &= \begin{pmatrix} \mathbf{1}^\top \\ \mathbf{X}^\top \end{pmatrix} \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix} \\
&= \begin{pmatrix} \mathbf{1}^\top \mathbf{1} & \mathbf{1}^\top \mathbf{X} \\ \mathbf{X}^\top \mathbf{1} & \mathbf{X}^\top \mathbf{X} \end{pmatrix} \\
&= \begin{pmatrix} n & \mathbf{1}^\top \mathbf{X} \\ \mathbf{X}^\top \mathbf{1} & \mathbf{X}^\top \mathbf{X} \end{pmatrix}.
\end{align*}
We have that
\begin{align*}
\mathbf{X}^\top \mathbf{X} - \mathbf{X}^\top \mathbf{1} \frac{1}{n} \mathbf{1}^\top \mathbf{X} &= \mathbf{X}^\top \mathbf{X} - (n \bar X) \frac{1}{n} (n \bar X^\top)\\
&= \mathbf{X}^\top \mathbf{X} - n \bar X \bar X^\top \\
&= n \hat\Sigma.
\end{align*}
As $\hat\Sigma$ is assumed invertible and $n \geq 1$, we can apply block matrix inversion, to obtain the desired.

#### 3. Show that if we fit a linear model with an intercept (which is denoted $\beta_0$) using the least squares estimator, then the variance of $\hat\beta_i$, conditionally on $\mathbf{X}$, for $i = 1, \dots, p$ is $\frac{\sigma^2}{n}(\hat\Sigma^{-1})_{ii}$

Write $\beta = (\beta_0, \beta_1, \dots, \beta_p)$ such that the first coordinate corresponds to the intercept. The belonging model matrix is $\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}$. 

Under the assumption of ***A1***, ***A2***, ***A4***, and that $\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}$ has full rank, Theorem 2.2 provides the variance of the ordinary least squares estimator
$$
\operatorname{Var}(\hat\beta \mid \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}) = \sigma^2 (\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}^\top \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix})^{-1}. 
$$

Using the result from the previous exercise, 
$$
\operatorname{Var}(\hat\beta \mid \begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}) = \sigma^2 \begin{pmatrix} \ast & \ast \\ \ast & \frac{1}{n}\hat \Sigma^{-1} \end{pmatrix}.
$$

Now it is just a matter of picking out the correct components of the covariance matrix, and we get, for each $i = 1, \dots, p$, that
$$
\operatorname{Var}(\hat\beta_i \mid \begin{pmatrix} \mathbf{1} & \mathbf{X}\end{pmatrix}) = \frac{\sigma^2}{n} (\hat \Sigma^{-1})_{ii}.
$$

Note that it does not matter whether we condition on $\mathbf{X}$ or $\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix}$ since $\sigma(\mathbf{X}) = \sigma(\begin{pmatrix} \mathbf{1} & \mathbf{X} \end{pmatrix})$. This proves the result. 

Let $\hat\Sigma_{-1}$ be the empirical covariance matrix with the first row and column removed. Furthermore, write $\hat\Sigma$ in block form as
$$
\hat \Sigma = \begin{pmatrix} \hat\sigma_1^2 & \hat\gamma^\top \\
\hat\gamma & \hat\Sigma_{-1}
\end{pmatrix}.
$$

#### 4. Use block inversion (again) to show that

$$
(\hat\Sigma_{11})^{-1} = \frac{1}{\hat\sigma_1^2 - \hat\gamma^\top (\hat\Sigma_{-1})^{-1} \hat\gamma}. 
$$


In the notation used in exercise 2, it holds that whenever $D$ is invertible, $P$ is invertible if and only if the Schur-complement of $D$ in $P$, $A-BD^{-1}C$ is invertible, and we have
$$
P^{-1} = \begin{pmatrix} A & B \\ C & D \end{pmatrix}^{-1} = \begin{pmatrix} (A-BD^{-1}C)^{-1} & \ast \\ \ast & \ast \end{pmatrix}.
$$
Since $\hat \Sigma$ is assumed invertible, it follows directly that

$$
(\hat\Sigma)^{-1}_{11} = \frac{1}{\hat\sigma_1^2 - \hat\gamma^\top (\hat\Sigma_{-1})^{-1} \hat\gamma}. 
$$

#### 5. Show that in the case $p = 2$, the formula above reduces to

$$
(\hat \Sigma^{-1})_{11} = \frac{1}{\hat\sigma_1^2 (1 - \widehat{\operatorname{corr}}(X_1, X_2)^2)}.
$$

Recall the definition of the correlation between two random variables (with finite variance) $X_1$ and $X_2$
$$
\operatorname{corr}(X_1, X_2) = \frac{\operatorname{cov}(X_1, X_2)}{\sqrt{\operatorname{Var}(X_1) \operatorname{Var}(X_2)}}
$$
whereby
$$
\operatorname{corr}(X_1, X_2)^2 = \frac{\operatorname{cov}(X_1, X_2)^2}{\operatorname{Var}(X_1) \operatorname{Var}(X_2)},
$$
with the plug-in estimator,
$$
\widehat{\operatorname{corr}}(X_1, X_2) = \frac{\widehat{\operatorname{cov}}(X_1, X_2)^2}{\hat{\sigma}_1^2 \hat{\sigma}_2^2}.
$$

If $p = 2$ then there are two (univariate) predictors $X_1$ and $X_2$, and $\hat{\Sigma}_{-1} = \hat{\sigma}_2^2$ and $\hat \gamma = \widehat{\operatorname{cov}}(X_1, X_2)$. The formula from the previous exercise now yields
\begin{align*}
(\hat\Sigma_{-1})_{11} &= \frac{1}{\hat\sigma_1^2 - \widehat{\operatorname{cov}}(X_1, X_2) \frac{1}{\hat{\sigma}^2_2}\widehat{\operatorname{cov}}(X_1, X_2)} \\
&= \frac{1}{\hat\sigma_1^2 (1 - \frac{\widehat{\operatorname{cov}}(X_1, X_2)^2}{\hat{\sigma}^2_2 \hat{\sigma}^1_2 } )} \\
&= \frac{1}{\hat\sigma_1^2 ( 1 - \widehat{\operatorname{corr}}(X_1, X_2)^2)},
\end{align*}
as desired. 

#### 6. Show that if $\mathbf{X}_1 - \mathbf{1} \bar X_1$ is perpendicular to the remaining columns in $\mathbf{X} - \mathbf{1} \bar X^\top$ , then

$$
(\hat\Sigma^{-1})_{11} = \frac{1}{\hat\sigma_1^2}.
$$

To distinguish columns from scalar entries, write the
$j$th column of the design matrix as $\mathbf X_{\cdot j}$ (as done in the exercise description). There are $p - 1$
remaining columns, indexed by $j = 2, \ldots, p$, so the orthogonality assumption
can be written as

$$
(\mathbf X_{\cdot 1} - \mathbf1 \bar X_1)^\top
(\mathbf X_{\cdot j}-\mathbf1 \bar X_j)=0,
\qquad j= 2, \ldots, p.
$$

The vector $\hat\gamma\in\mathbb R^{p-1}$ contains the
empirical covariances between the first predictor and the remaining predictors.
More precisely, for $j = 2, \ldots, p$,

$$
\hat\gamma_{j-1}
=\frac1n
(\mathbf X_{\cdot1}-\mathbf1\bar X_1)^\top
(\mathbf X_{\cdot j}-\mathbf1\bar X_j).
$$

Thus orthogonality implies $\hat\gamma=\mathbf0_{p-1}$. 

The result from exercise 4 now yields
\begin{align*}
(\hat\Sigma^{-1})_{11} = \frac{1}{\hat\sigma_1^2 - \hat\gamma^\top (\hat\Sigma_{-1})^{-1} \hat\gamma} = \frac{1}{\hat\sigma_1^2}.
\end{align*}


#### 7. Interpret the results obtained in this exercise in terms of how the empirical variance of and empirical correlation between predictor variables affect the variances of the estimators $\hat \beta_i$ for $i = 1, \dots, p$.

For $p = 2$, recall that
$$
\operatorname{Var}(\hat \beta_1 \mid \mathbf{X}) = \frac{\sigma^2}{n} \frac{1}{\hat\sigma_1^2(1 - \widehat{\operatorname{corr}}(X_1, X_2)^2)}.
$$

This is decreasing in if $\hat\sigma_1^2$ increases, and it blows up whenever $\widehat{\operatorname{corr}}(X_1, X_2)$ approaches $1$ or $-1$. Thus, the (conditional) variance of $\hat \beta_1$ is very large if the empirical variance of $X_1$ is very small or if the predictors are very correlated (both positively and negatively). 

For $p > 2$, we see from exercise 6 that the (conditional) variance of $\hat\beta_1$ is minimized when $\mathbf{X}_{\cdot 1} - \mathbf{1} \bar X_1$ is orthogonal to the other centered predictors, because $\hat\Sigma_{-1}$ is positive semi-definite and thus $\hat\gamma^\top (\hat\Sigma_{-1})^{-1} \hat\gamma \geq 0$ with equality when the orthogonality assumption holds. 

For the remaining coordinates of $\beta$, we can simply permute both the coefficient vector $\beta$ and the model matrix $\mathbf{X}$ such that the variance of $\hat\beta_j$ is minimized when $\mathbf{X}_{\cdot j} - \mathbf{1} \bar X_j$ is orthogonal to the other centered predictors. 

## Exercise 3 - 3.1

We load the dataset from the `RwR` package and discard the unemployed from the dataset. As per the exercise hint, we also collapse the respondents with 3 or more children into one group. We do this by adding people with more than 3 children to the 3 children group. We choose to retain children as numeric, rather than creating a factor.

```{r}
data(wage, package = "RwR")
wage_employed <- as_tibble(wage) |> filter(hours>0) |> mutate(children = pmin(3,children))
wage_unemployed <- as_tibble(wage) |> filter(hours==0) |> mutate(children = pmin(3,children))

```

#### 1. Fit and report a linear, additive model of the mean wage given some of the other variables using data for the employed respondents only.

We fit a linear additive model of mean using all other covariates and report parameter estimates

```{r}
lin_mod <- lm(wage ~ . , data = wage_employed)
tidy(lin_mod) |>
  knitr::kable()
```
We see that (perhaps unsurprisingly) there is a positive association between hours worked and wage. There is also a positive association both between the number of children and wage, as well as age and wage. The expected wage is lowest for people with who graduated high-school but not college, followed by people without a high-school diploma and largest for people with a college degree. We can also use the `drop1` function to drop one of the predictors.

```{r}
drop1(lin_mod, test = "F") |> 
  tidy() |> 
  arrange(p.value, desc(statistic)) |>
  filter(term != "<none>") |>
  select(-AIC) |>
  knitr::kable()
```

It seems like everything except for education is a fairly strong predictor.

#### 2.

We do model diagnostics
```{r}
#| message: false
#| warning: false
data_diag1 <- augment(lin_mod)
p1 <- ggplot(data_diag1, aes(.fitted, .std.resid)) +
  geom_point() + geom_smooth(formula = y ~ x) +
  xlab("Fitted Values") +
  ylab("Standardized Residuals")

p2 <- ggplot(data_diag1, aes(sample = .std.resid)) +
  geom_qq() + geom_abline() +
  xlab("Theoretical quantiles") +
  ylab("")

p3 <- ggplot(data_diag1, aes(hours, .std.resid)) +
  geom_point() + geom_smooth(formula = y ~ x) +
  xlab("Hours") +
  ylab("Standardized Residuals")

p4 <- ggplot(data_diag1, aes(age, .std.resid)) +
  geom_point() + geom_smooth(formula = y ~ x) +
  xlab("Age") +
  ylab("Standardized Residuals")


grid.arrange(p1, p2, p3, p4, ncol = 2)
```

The mean value structure is not entirely captured for both the small and the large fitted values. This could be the case, as there seems to be a non-linear effect of wage, and to a lesser extent hours worked. Overall, the qq-plot looks ok.

#### 3.

We will here focus on modelling the seemingly non-linear effect of age better. The above residual plot suggests that we can add $\text{age}^2$ and $\text{age}^3$ to better capture the non-linear effect. Beware that this type of polynomial regression typically extrapolates poorly outside the data range. In chapter 7, we will see that splines are typically more appropriate for handling non-linearities.

```{r}
#| message: false
#| warning: false

cubic_mod <- lm(wage ~ age + I(age^2) + I(age^3) + hours + educ + children , data = wage_employed)


data_diag_cubic <- augment(cubic_mod)
p1 <- ggplot(data_diag_cubic, aes(.fitted, .std.resid)) +
  geom_point() + geom_smooth(formula = y ~ x) +
  xlab("Fitted Values") +
  ylab("Standardized Residuals")

p2 <- ggplot(data_diag_cubic, aes(sample = .std.resid)) +
  geom_qq() + geom_abline() +
  xlab("Theoretical quantiles") +
  ylab("")

p3 <- ggplot(data_diag_cubic, aes(hours, .std.resid)) +
  geom_point() + geom_smooth(formula = y ~ x) +
  xlab("Hours") +
  ylab("Standardized Residuals")

p4 <- ggplot(data_diag_cubic, aes(age, .std.resid)) +
  geom_point() + geom_smooth(formula = y ~ x) +
  xlab("Age") +
  ylab("Standardized Residuals")


grid.arrange(p1, p2, p3, p4, ncol = 2)

```

Since the linear model is nested in the cubic model, we can compare the two models via an $F$-test.
```{r}
anova(cubic_mod,lin_mod) |>
  knitr::kable()
```
We reject the null hypothesis of a linear effect in age against the model with a cubic effect of age. The cubic and quadratic term on age, makes interpreting the estimated parameters difficult. Instead, we opt for plotting the predictions of the model for varying levels of age, for an individual who works the average number of hours, has two children and a high-school diploma. As the other than age covariates enter the model linearly, varying these simply correspond to additive shifts in the predicted values.

```{r}
age_seq <- seq(min(wage_employed$age), max(wage_employed$age), by = 0.01)
df_pred <- tibble(age = age_seq, educ = factor(2), hours = mean(wage$hours), children = 2) |>
  mutate(predicted = predict(cubic_mod, newdata = cur_data()))

ggplot(df_pred) +
  geom_line(aes(age, predicted), color = "blue") +
  xlab("Age") + 
  ylab("Predicted Wage")

```
Note that this is just one way of visualing the model, and there are many other possible choices.

#### 4.

The model we have chosen is actually not particularly well-suited for predicting the wage distribution for the unemployed respondents, should they get a job. The reasons for this, is that we include hours worked in the model, which will also increase (from 0) if the respondents are no longer unemployed. We here take a pragmatic approach, and assume that if the respondents finds employment they will work the average number of hours that the employed people work. Be aware that this is an assumption that can not be verified from data. We plot (estimated) densities for the predicted wage distributions among the unemployed and the actual wage distribution in employed group. We cap the predicted wage at 0. 

```{r}
mean_hrs_employed <- mean(wage_employed$hours)

dist_unemployed <- wage_unemployed |> 
  mutate(hours = mean_hrs_employed) |> 
  mutate(pred_wage = pmax(0,predict(cubic_mod, cur_data())), Employed = FALSE) |>
  select(wage = pred_wage,Employed)

dist_employed <- wage_employed |>
  mutate(Employed = TRUE) |>
  select(wage,Employed)

bind_rows(dist_employed, dist_unemployed) |>
  ggplot(aes(wage, color = Employed)) + 
  geom_density()


```
We find a much smaller spread in the predicted wages compared to the realised wages, and that we in expect that the unemployed would earn more on average than the ones who are currently employed. This may not seem particularly realistic, and may just be an artefact of the fact that we used the mean hours worked in the employed group for the hours worked by the unemployed. We could also fit the same model without hours worked,

```{r}
cubic_mod_no_hrs <- lm(wage ~ age + I(age^2) + I(age^3) + educ + children , data = wage_employed)

dist_unemployed <- wage_unemployed |> 
  mutate(pred_wage = pmax(0,predict(cubic_mod_no_hrs, cur_data())), Employed = FALSE) |>
  select(wage = pred_wage,Employed)

dist_employed <- wage_employed |>
  mutate(Employed = TRUE) |>
  select(wage,Employed)

bind_rows(dist_employed, dist_unemployed) |>
  ggplot(aes(wage, color = Employed)) + 
  geom_density()


```
And now we obtain the opposite conclusion w.r.t. mean wage.





