---
title: "Week 1: Monday Solutions"
author: "Kasper Niss, Asbjørn Risom"
---

## Exercise 1 - Install RwR package and reproduce results

#### RwR package

The [`RwR` package](https://github.com/nielsrhansen/RwR/tree/main/RwR_package) contains the datasets and case studies used in *Regression with R*. Install it directly from GitHub:

We first install `remotes` from CRAN and then use it to install `RwR` from GitHub:

```{r, message = FALSE, results = 'hide'}
# Install remotes from CRAN
install.packages("remotes")

# Install RwR from GitHub
remotes::install_github(
  "nielsrhansen/RwR/RwR_package"
)
```

Load the installed (and other relevant) package(s):

```{r}
#| echo: true
#| message: false
#| warning: false
library(RwR)
library(ggplot2)
library(tibble)
library(broom)
library(splines)
library(gridExtra)
library(R.utils)
theme_set(theme_bw()) # plotting theme is always theme_bw()
```

#### Reproducing results from Section 2.2 and 2.4 in RwR
We load the `claims` dataset and take a look at it. 

```{r}
data(claims, package = "RwR")
tibble(claims) |> head()
```

Recall the trade group encoding:
1 = residences
2 = ordinary industry
3 = hotel and restaurant
4 = production industry.

We plot the histogram of claim size and insurance (on log scale)

```{r}
grid.arrange(
  ggplot(data = claims, aes(x = claims)) + 
    geom_histogram(color = "white") + 
    scale_x_log10(),
  ggplot(data = claims, aes(x = sum)) + 
    geom_histogram(color = "white") + 
    scale_x_log10()
)
```

Note that we use the default number of bins so the histogram might not look exactly as the one in RwR book. 

Next we plot claims against sum  (on log-log-scale) and fit a linear regression model

```{r}
p0 <- ggplot(
    data = claims, 
    aes(sum, claims)
) + 
    scale_x_log10(
        "Insurance sum (DKK)", 
        breaks = 10^c(6, 7, 8), 
        labels = c("1M", "10M", "100M")
    ) +
    scale_y_log10(
        "Claim size (DKK)", 
        breaks = 10^c(2, 4, 6, 8), 
        labels = c("100", "10K", "1M", "100M")
    ) +
    geom_point(alpha = 0.2)

p <- p0 +
    geom_smooth(
        method = "lm", 
        linewidth = 1, 
        se = FALSE
    )

p
```

We can facet on each trade group, that is, we plot the scatter plot for each trade group 

```{r}
p + facet_wrap(~ grp, ncol = 4)
```

We now fit a linear model with `log(claims)` as response and `log(sum)` as the predictor

```{r}
claims_lm <- lm(
    log(claims) ~ log(sum), 
    data = claims
)
coefficients(claims_lm)
```

corresponding to the fitted line in the scatter plot where we don't split according to trade group. 

We also look at the model matrix

```{r}
model.matrix(claims_lm)[781:784, ]
```

Next we fit a model with seperate intercept for each trade group

```{r}
claims_lm_add <- lm(
    log(claims) ~ log(sum) + grp,
    data = claims
)
coefficients(claims_lm_add)
```

and (parts of) the model matrix is

```{r}
model.matrix(claims_lm_add)[781:784, ]
```

An alternative parametrization as well

```{r}
claims_lm_add <- lm(
    log(claims) ~ log(sum) + grp, 
    data = claims,
    contrasts = list(grp = "contr.sum")
)
coefficients(claims_lm_add)
```

Now we fit the interaction model with intercations between `log(sum)` and the trade group
```{r}
claims_lm_int <- lm(log(claims) ~ log(sum) * grp, data = claims)
coefficients(claims_lm_int) |> print(width = 120)
```

```{r}
model.matrix(claims_lm_int)[781:784, ] |> print(width = 120)
```

This model corresponds to the figure where we faceted on each trade group. 

## Exercise 2 - 2.1

We simulate a dataset to work with and visualize it.

```{r}
set.seed(04022016)
factor_labels <- rep(c("A", "B", "C"), each = 50)
formula_data <- tibble(
    x =  rnorm(150, 0),
    z = factor(sample(factor_labels)),
    mu = (z == "A") + cos(x) + 0.5 * (z == "B") * sin(x),
    y = rnorm(150, mu, 0.5)
)

p0 <- ggplot(data = formula_data, aes(x = x, y = y)) +
  geom_point()
p0
```

#### 1. Fit a linear regression model using `lm(y ~ x, formula_data)` where `y` is the outcome variable and `x` is the predictor variable. Investigate the residuals. Does the model fit the data?

We fit the linear model

```{r}
linearfit <- lm(formula = y ~ x, data = formula_data)
```

and inspect the fit

```{r, message = FALSE}
linearfit
p0 + geom_smooth(method = "lm", formula = y ~ x)
```

and it looks like a straight line is a bad fit for this data. 

To plot the residuals we use `augment` 

```{r}
linearfit_diag <- augment(linearfit) # diag = diagnostics

gridExtra::grid.arrange(
  ggplot(linearfit_diag, aes(.fitted, .resid)) +
    geom_point(alpha = 0.2) +
    geom_smooth(),

  ggplot(linearfit_diag, aes(.resid)) +
    geom_histogram(bins = 40),

  ggplot(linearfit_diag, aes(sample = .std.resid)) +
    geom_qq() +
    geom_qq_line(),

  ncol = 3
)
```

Residuals plotted against fitted values (plot to left) shows the mean value assumption (A1) seems to be violated: the mean of the residuals are not zero. 

#### 2. How do you get rid of the intercept via the formula specification?

We get rid of the intercept by adding a `-1`to the model fit:

```{r}
no_intercept_linearfit <- lm(y ~ x - 1, data = formula_data)

no_intercept_linearfit |> model.matrix() |> head()

p0 + 
  geom_smooth(method = "lm", formula = y ~ x - 1)
```

#### 3. How can you check (numerically) that the two model matrices above span the same column space?

Recall from (2.12) in the notes that the column space of $\mathbf{X}'$ is contained in the column space of $\mathbf{X}$ iff there exists a matrix $C$ such that
$$
\mathbf{X}\mathbf{C} = \mathbf{X}'
$$
This is equivalent to 
\begin{equation*}
\mathbf{X}^{\top}\mathbf{X}\mathbf{C} = \mathbf{X}^{\top}\mathbf{X}'
\end{equation*}
Note that if $X$ has full rank, the matrix $C$ is given by
$$
\mathbf{C} = \left(\mathbf{X}^{\top}\mathbf{X}\right)^{-1} \mathbf{X}^{\top}\mathbf{X}',
$$
which explains the result from Exercise 2.3. In the any case, we can solve linear equations via the `` R`` funcition ``solve`` function.
```{r}
X1 <- model.matrix(y ~ x + z, formula_data)

contrasts(formula_data$z) <- contr.treatment(levels(formula_data$z), base = 3)
X2 <- model.matrix(y ~ x + z, formula_data)

check_equality_of_col_space <- function(X1,X2){
  C_mat <- solve(t(X2) %*% X2, t(X2) %*% X1) 
  res_sq1 <- sum((X2 %*% C_mat - X1)^2)
  
  C_mat <- solve(t(X1) %*% X1, t(X1) %*% X2)
  res_sq2 <- sum((X1 %*% C_mat - X2)^2)
  
  isZero(res_sq1) && isZero(res_sq2)
}
check_equality_of_col_space(X1,X2)

```


#### 4. Are the two models below identical? That is, do the columns of the two model matrices span the same column space?

The two matrices are
```{r}
X1 <- model.matrix(y ~ z + x : z, formula_data)
X1 |> head()

X2 <- model.matrix(y ~ x * z, formula_data)
X2 |> head()
```

We use the function from above to check if the column spaces are equal

```{r}
check_equality_of_col_space(X1,X2)
```

#### 5. Run the following code and explain what it returns.


```{r}
model.matrix(y ~ cos(x) + z * sin(x), formula_data) |>
    head()
```

The code creates the model matrix for the model with predictor `cos(x)`, the factors `z`, and the predictors `sin(x)`. 

Recall that `z * sin(x) = z + sin(x) + z:sin(x)` where the latter is the interaction term. That is why we see `cos(x), zA, zB, sin(x)` and the interactions between `zA` and `zB` with `sin(x)`. Recall that the baseline factor is `C`. There is also an intercept. 

We allow for different intercepts in the group with the factors and the coefficient of `sin(x)` can change for each group. 

#### 6. Fit the model specified by the formula ``y ~ cos(x) + z * sin(x)`` to the simulated data, explain the parameter estimates you get and reproduce Figure 2.7.

```{r}
interact_model <- lm(y ~ cos(x) + z * sin(x), data = formula_data)
```

The parameter estimates can be found in

```{r}
summary(interact_model)
```

Recall that `C` is the baseline so that `(Intercept)` and `sin(x)` describe this baseline. `zA` and `zB` are intercept differences from `C` and `zA:sin(x)` and `zB:sin(x)` are the changes in the slope of `sin(x)` for each group, relative to `C`.  The coefficient in front of `cos(x)` is the common coefficient of `cos(x)` for all three groups

The fitted group-specific means are approximately (here the coffeciients are rounded)
$$
\begin{aligned}
\widehat\mu_C(x) &= -0.048 + 1.046\cos(x) + 0.128\sin(x),\\
\widehat\mu_A(x) &= 0.979 + 1.046\cos(x) + 0.055\sin(x),\\
\widehat\mu_B(x) &= 0.069 + 1.046\cos(x) + 0.524\sin(x).
\end{aligned}
$$

Recall that data is simulated with a coefficient of 1 in front of `cos(x)`, and the sine coefficient for group `B` is 0.5 -- both are recovered quite well. The coefficient for group `A` and `C` are 0, with both estimates reasonably close to 0. 

Finally, we reconstruct the plot. 

First we define append the fitted values from the model fit

```{r}
formula_data$fitted <- predict(interact_model, newdata = formula_data)
```

Then we construct the true mean-value functions

```{r}
mean_function <- function(x, z) {
  (z == "A") + cos(x) + 0.5 * (z == "B") * sin(x)
}
```


```{r}
ggplot(formula_data, aes(x = x, y = y, color = z)) +
  geom_point() +
  geom_line(aes(y = fitted)) +
  geom_function(
    fun = \(x) mean_function(x, "A"),
    aes(color = "A"),
    linetype = 2
  ) +
  geom_function(
    fun = \(x) mean_function(x, "B"),
    aes(color = "B"),
    linetype = 2
  ) +
  geom_function(
    fun = \(x) mean_function(x, "C"),
    aes(color = "C"),
    linetype = 2
  )
```


Finally also the residual plot, which is easy since we have augmented `formula_data` with the fitted values. We start by calculating the residuals

```{r}
formula_data$resid <- formula_data$y - formula_data$fitted
```

and then we plot

```{r}
ggplot(data = formula_data, aes(x = fitted, y = resid)) + 
  geom_point(aes(color = z)) + 
  geom_smooth()
```

