---
title: "Week 5: Monday Solutions"
author: "Kasper Niss & Asbjørn Risom"
format:
  html:
    include-in-header:
      text: |
        <script>
        window.MathJax = {
          tex: {
            macros: {
              norm: ["\\left\\lVert #1 \\right\\rVert", 1],
            }
          }
        };
        </script>
---



```{r, include=FALSE}
library(RwR)
library(dplyr)
library(tidyverse)
library(RwR)
library(ggplot2)
library(tibble)
library(broom)
library(splines)
library(gridExtra)
theme_set(theme_bw())
```

## Exercise 8.2 

Recall that a linear estimator is of the form 
$$
\mathbf{\hat\mu} = \mathbf{S} \mathbf{Y}
$$ 
for an $\mathbf{X}$ dependent matrix $\mathbf{S}$. Here, $\mu = \mathbb{E}[\mathbf{Y} \mid \mathbf{X}] \in \mathbb{R}^n$ is the vector of conditional expectations of the outcome given the predictors.

#### 1. Show that the penalized and weighted least squares estimator, solving the normal equation in Theorem 2.1, is linear and identify the matrix $\mathbf{S}$.

Note that the question is slightly ambigious. It is unclear whether it is the estimated coefficient vector or the fitted mean we need to show is linear.

In fact, they are both!

Under the assumption of uniqueness in Theorem 2.1 -- either $\mathbf{X}$ has full column rank or $\Omega$ is positive definite -- the penalized and weighted least squares estimator of the coefficient vector is $$
\hat\beta_\lambda = (\mathbf{X}^\top \mathbf{W} \mathbf{X} + \Omega)^{-1} \mathbf{X}^\top \mathbf{W} \mathbf{Y} = \mathbf{B} \mathbf{Y},
$$ where $\mathbf{B} = (\mathbf{X}^\top \mathbf{W} \mathbf{X} + \Omega)^{-1} \mathbf{X}^\top \mathbf{W}$ showing that $\hat\beta_\lambda$ is linear.

The induced fitted mean estimator is $\hat\mu_\lambda = \mathbf{X}\hat\beta_\lambda$, and we see that $$
\hat\mu_\lambda = \mathbf{X} \mathbf{B} \mathbf{Y} = \mathbf{S} \mathbf{Y},
$$ where $\mathbf{S} = \mathbf{X}(\mathbf{X}^\top \mathbf{W} \mathbf{X} + \Omega)^{-1} \mathbf{X}^\top \mathbf{W}$.

This shows that both $\hat\beta_\lambda$ and $\hat\mu_\lambda$ are linear.

For the rest of this exercise we assume that the weight matrix is $\mathbf{W} = \mathbf{I}$ and the penalty matrix is diagonal and of the form $$
\Omega = \lambda
\begin{pmatrix}
\omega_1 & 0 & \dots & 0 \\
0 & \omega_2 & \dots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \dots & \omega_p
\end{pmatrix}
 = \lambda \operatorname{diag}(\omega) = \lambda \operatorname{diag}(\omega_1, \dots, \omega_p),
$$ where $\lambda > 0$. You should think about $\lambda$ as a tuning parameter that we want to optimize over, while $\omega_1, \dots, \omega_p \geq 0$ are fixed penalty weights determining the relative distribution of penalization on the different parameters. We also let $\hat\beta$ and $\hat\mu$ denote the (unpenalized) least squares estimator of $\beta$ and $\mu$ respectively.

#### 2. Show that if the model matrix $\mathbf{X}$ has orthogonal columns, i.e. $\mathbf{X}^\top \mathbf{X}$ is diagonal, then with

$$
\gamma_j = \sum_{i = 1}^{n} X_{ij}^2 = \norm{\mathbf{X}_{\cdot, j}}^2
$$

the squared 2-norm of the $j$-th column of $\mathbf{X}$, the penalized least squares estimator, $\hat\beta_\lambda$,is given by $$
\hat\beta_{\lambda,j} = \frac{\gamma_j}{\gamma_j + \lambda \omega_j}\hat\beta_j,
$$

and $$
\hat\mu_\lambda = \mathbf{X} \hat\beta_\lambda.
$$ Show also that the effective degrees of freedom is $$
\operatorname{df}_\lambda = \sum_{j = 1}^{p} \frac{\gamma_j}{\gamma_j + \lambda \omega_j}.
$$

Recall the formula for $\hat\beta_\lambda$ from above, and note that $$
\hat\beta_\lambda = (\mathbf{X}^\top \mathbf{W} \mathbf{X} + \Omega)^{-1} \mathbf{X}^\top \mathbf{W} \mathbf{Y} = (\mathbf{X}^\top \mathbf{X} + \Omega)^{-1}\mathbf{X}^\top \mathbf{Y}
$$ because $\mathbf{W} = \mathbf{I}$. By the ortogonality assumption, the columns in $\mathbf{X}$ satisfy $$
\mathbf{X}_{\cdot, j}^\top \mathbf{X}_{\cdot,k} = 0, \quad j \neq k,
$$ and for $j = k$, $$
\mathbf{X}_{\cdot, j}^\top \mathbf{X}_{\cdot, j} = \norm{\mathbf{X}_{\cdot, j}}^2 = \gamma_j.
$$ That is, $$
\mathbf{X}^\top \mathbf{X} = \operatorname{diag}(\gamma) = \operatorname{diag}(\gamma_1, \dots, \gamma_p),
$$ a diagonal matix with $\gamma_1, \dots, \gamma_p$ on the diagonal.

The penalized least squares estimator (which is not weigthed because $\mathbf{W} = \mathbf{I}$) then reduces to \begin{align*}
\hat\beta_\lambda &= (\mathbf{X}^\top \mathbf{X} + \Omega)^{-1}\mathbf{X}^\top \mathbf{Y} \\
 &= (\operatorname{diag}(\gamma) + \lambda \operatorname{diag}(\omega))^{-1} \mathbf{X}^\top \mathbf{Y} \\
&= (\operatorname{diag}(\gamma + \lambda \omega))^{-1}\mathbf{X}^\top \mathbf{Y} \\
&= \operatorname{diag}\left(\frac{1}{\gamma + \lambda \omega}\right) \mathbf{X}^\top \mathbf{Y}. 
\end{align*}

Similarly, the unpenalized estimator, $\hat\beta$, under orthogonality, is given as $$
\hat\beta = (\mathbf{X}^\top \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{Y} = \operatorname{diag}\left(\frac{1}{\gamma}\right) \mathbf{X}^\top \mathbf{Y},
$$ and combining the two, we see that \begin{align*}
\hat\beta_\lambda &= \operatorname{diag}\left(\frac{1}{\gamma + \lambda \omega}\right) \mathbf{X}^\top \mathbf{Y} \\
&= \operatorname{diag}\left(\frac{\gamma}{\gamma + \lambda \omega}\right) \operatorname{diag}\left(\frac{1}{\gamma}\right) \mathbf{X}^\top \mathbf{Y} \\
&= \operatorname{diag}\left(\frac{\gamma}{\gamma + \lambda \omega}\right) \hat\beta,
\end{align*} and the result follows by choosing the $j$-th coordinate.

For the degrees of freedom, we use example 8.7, which applies under the canonical link being the identity (i.e. a linear model), and also assuming A2 and A4, $$
\operatorname{df}_\lambda = \operatorname{trace}(\mathbf{S}) = \operatorname{trace}(\mathbf{X}(\mathbf{X}^\top \mathbf{W} \mathbf{X} + \Omega)^{-1} \mathbf{X}^\top \mathbf{W}).
$$ Recall again that $\mathbf{W} = \mathbf{I}$, $\mathbf{X}^\top \mathbf{X} = \operatorname{diag}(\gamma)$, and that $\Omega = \lambda \operatorname{diag}(\omega)$, such that $$
\mathbf{S} = \mathbf{X}\operatorname{diag}\left(\frac{1}{\gamma + \lambda \omega}\right) \mathbf{X}^\top.
$$ Now using the cyclic property of the $\operatorname{trace}$ operator, we find that \begin{align*}
\operatorname{df}_\lambda &= \operatorname{trace}(\mathbf{S}) \\
&= \operatorname{trace}\left(\mathbf{X}\operatorname{diag}\left(\frac{1}{\gamma + \lambda \omega}\right) \mathbf{X}^\top\right) \\ 
&= \operatorname{trace}\left(\operatorname{diag}\left(\frac{1}{\gamma + \lambda \omega}\right) \mathbf{X}^\top \mathbf{X}\right) \\
&= \operatorname{trace}\left(\operatorname{diag}\left(\frac{1}{\gamma + \lambda \omega}\right) \operatorname{diag}(\gamma) \right) \\ 
&= \operatorname{trace}\left(\operatorname{diag}\left( \frac{\gamma}{\gamma + \lambda \omega}\right)\right) \\
&= \sum_{j = 1}^{p} \frac{\gamma_j}{\gamma_j + \lambda \omega_j}.
\end{align*}

#### 3. Show that if $\omega_1 = \dots = \omega_p = 1$, then, irrespectively of whether $\mathbf{X}$ has orthogonal columns, the effective degrees of freedom is

$$
\operatorname{df}_\lambda = \sum_{j = 1}^{p} \frac{\gamma_j}{\gamma_j + \lambda},
$$ where $\gamma_1, \dots, \gamma_p \geq 0$ are the eigenvalues of $\mathbf{X}^\top \mathbf{X}$.

Since $\omega_1 = \dots = \omega_p = 1$, $\Omega = \lambda \mathbf{I}$.

Because $\mathbf{X}^\top \mathbf{X}$ is real and symmetric, we can eigenvalue decompose it as $\mathbf{X}^\top \mathbf{X} = \mathbf{Q} \mathbf{V} \mathbf{Q}^\top$, where $\mathbf{Q}$ is an orthonormal matrix (i.e. $\mathbf{Q}^\top \mathbf{Q} = \mathbf{I}$ and $\mathbf{Q}^{-1} = \mathbf{Q}^\top$) consisting of the eigenvectors of $\mathbf{X}^\top \mathbf{X}$ and $\mathbf{V}$ is a diagonal matrix consisting of the eigenvalues of $\mathbf{X}^\top \mathbf{X}$, that is $\mathbf{V} = \operatorname{diag}(\gamma) = \operatorname{diag}(\gamma_1, \dots, \gamma_p)$.

When $\omega = \mathbf{1}$, $\Omega = \lambda \mathbf{I}$, and $\mathbf{S} = \mathbf{X}(\mathbf{X}^\top\mathbf{X} + \lambda \mathbf{I})^{-1} \mathbf{X}^\top$.

Using the eigendecomposition of $\mathbf{X}^\top\mathbf{X}$, we get $$
\mathbf{X}^\top\mathbf{X} + \lambda \mathbf{I} = \mathbf{Q}(\mathbf{V} + \lambda \mathbf{I}) \mathbf{Q}^\top.
$$ The orthogonality of $\mathbf{Q}$ gives $$
(\mathbf{X}^\top\mathbf{X} + \lambda \mathbf{I})^{-1} = \mathbf{Q}^{-\top}(\mathbf{V} + \lambda \mathbf{I})^{-1} \mathbf{Q}^{-1} = \mathbf{Q} (\mathbf{V} + \lambda \mathbf{I})^{-1} \mathbf{Q}^\top.
$$ The effective degrees of freedom are then \begin{align*}
\operatorname{df}_\lambda &= \operatorname{trace}(\mathbf{S}) \\
&= \operatorname{trace}\left(\mathbf{X}(\mathbf{X}^\top\mathbf{X} + \lambda \mathbf{I})^{-1} \mathbf{X}^\top\right) \\
&= \operatorname{trace}\left((\mathbf{X}^\top\mathbf{X} + \lambda \mathbf{I})^{-1} \mathbf{X}^\top\mathbf{X}\right) \\
&= \operatorname{trace}\left(\mathbf{Q} (\mathbf{V} + \lambda \mathbf{I})^{-1} \mathbf{Q}^\top \mathbf{Q}\mathbf{V}\mathbf{Q}^\top \right) \\
&= \operatorname{trace}\left((\mathbf{V} + \lambda \mathbf{I})^{-1} \mathbf{V}\right) \\
&= \operatorname{trace}\left(\operatorname{diag}\left(\frac{1}{\gamma + \lambda}\right)\operatorname{diag}(\gamma)\right) \\
&= \sum_{j = 1}^{p} \frac{\gamma_j}{\gamma_j + \lambda}.
\end{align*}

For the remaining part of this exercise we consider the bike rental data introduced in Section 8.2. Using the `poly()` function we construct a model matrix with $p = 7$ orthogonal columns, corresponding to expanding the regression function using orthogonal polynomials of degree $6$.

```{r}
data(bike, package = "RwR")
X <- model.matrix(~ poly(Temperature, degrees = 6), data = bike)
beta_hat <- lm.fit(X, bike$Count) |> coefficients()
```

The least squares estimate is computed above as `beta_hat`, and the corresponding fitted values are shown in Figure 8.12 (or below). For the following questions we take $\omega_1 = \omega_2 = 0$ and $\omega_3 = \dots \omega_7 = 1$.

```{r}
ggplot(bike, aes(Temperature, Count)) + 
  geom_point() + 
  geom_line(aes(y = X %*% beta_hat))
```

#### 4. Check numerically that `X` (approximately) has *orthogonal* columns.

We check that the product $\mathbf{X}^\top \mathbf{X}$ is approximately diagonal

```{r}
XtX <- t(X) %*% X

off.diagonal <- XtX
diag(off.diagonal) <- 0 #set the diagonal in X^top X to be 0
max(abs(off.diagonal)) #find the largest value on the off diagonal
```

It is small, thus $\mathbf{X}^\top \mathbf{X}$ is approximately diagonal.

#### 5. Compute an estimate of $\sigma^2$ based on the least squares estimate $\hat\beta$ of $\beta$.

Recall (see e.g. equation (2.9) and Theorem 2.2) that an estimator of $\sigma^2$ is $\hat\sigma^2 = \frac{1}{n - p}\norm{Y - \mathbf{X} \hat\beta}^2$, where $n = `{r} nrow(bike)`$ and $p = `{r} length(beta_hat)`$ that is the number of datapoints and covariates, respectively.

```{r}
n <- nrow(bike)
p <- length(beta_hat)

resid <- bike$Count - X %*% beta_hat

sigma.hat.sq <- 1 / (n - p) * sum(resid^2)
sigma.hat.sq
```

We could also use the `lm` function

```{r}
fit <- lm(Count ~ poly(Temperature, degree = 6), data = bike)
sigma(fit)^2
```

#### 6. Implement the computation of $\hat\mu_\lambda$ and $\operatorname{df}_\lambda$ for a sequence of values of $\lambda$.

Note that $\lambda = 0$ gives least squares and large values of $\lambda$ effectively make $\hat\beta_\lambda = (\hat\beta_0, \hat\beta_1, 0, \dots, 0)^\top$.

Recall that $\omega = (0, 0, 1, \dots, 1)^\top \in \mathbb{R}^7$

```{r}
omega <- c(0, 0, rep(1, ncol(X) - 2))
```

Because $\mathbf{X}$ has ortogonal columns, see question 4, we can use the result from question 2 to calculate both $\hat\mu_\lambda$ and $\operatorname{df}_\lambda$.

First we find the vector $\gamma$ as the squared 2-norm of the $p = `{r} ncol(X)`$ columns of $\mathbf{X}$ and then we create functions to compute $\hat\mu_\lambda$ and $\operatorname{df}_\lambda$

```{r}
gamma <- colSums(X^2) # finds the squared norm

mu.hat.lambda <- function(lambda) {
  shrinkage <- gamma / (gamma + lambda * omega)
  beta.hat.lambda <- shrinkage * beta_hat # beta_hat is not penalized

  drop(X %*% beta.hat.lambda)
}

df.lambda <- function(lambda) {
  shrinkage <- gamma / (gamma + lambda * omega)

  sum(shrinkage)
}
```

We try it for various $\lambda$s

```{r}
lambda.values <- c(0, 10^seq(-4, 4, length.out = 100))

mu.hat.path <- sapply(lambda.values, mu.hat.lambda)
df.path <- sapply(lambda.values, df.lambda)

lambda.results <- data.frame(
  lambda = lambda.values,
  df = df.path
)
```

Since $\lambda = 0$ gives the least squares estimator we expect $\operatorname{df}_0 = p = 7$, and because the first two coefficients have penalty weights equal to $0$, ($\omega_1 = \omega_2 = 0$), we also expect $\operatorname{df}_\lambda \to 2$ as $\lambda \to \infty$. 2 MAKES SENSE BECAUSE ONLY TWO COMPONENTS IN BETA HAT BUT WHAT HAS THIS TO DO WITH 0 WEIGHTS??

```{r}
c(
  df.at.lambda.0 = df.lambda(0),
  df.at.large.lambda = df.lambda(1e8)
)
```

We also plot the effective degrees of freedom against the values of $\lambda$

```{r}
#| warning: false
#| message: false

lambda.results |>
  filter(lambda > 0) |>
  ggplot(aes(x = lambda, y = df)) +
  geom_point(size = 0.2) + 
  geom_line() +
  scale_x_log10() +
  labs(
  x = expression(log[10](lambda)),
  y = expression(df[lambda])
)
```

#### 7. Compute $\operatorname{err}$ for the sequence of $\lambda$-values and compute then a sequence of estimates of the in-sample error. Plot the in-sample error against the effective degrees of freedom. What is the best model according to the in-sample error?

Recall that we are using the linear model so the error is using least squares, i.e. $$
\operatorname{err}_\lambda = \frac{1}{n} \norm{\mathbf{Y} - \hat\mu_\lambda}^2
$$

```{r}
err <- colMeans((bike$Count - mu.hat.path)^2)

lambda.results <- lambda.results |>
  mutate(
    err = err
  )
```

We can plot $\operatorname{err}$ as a function of the effective degrees of freedom

```{r}
err.plot <- ggplot(
  lambda.results,
  aes(x = df, y = err)
) +
  geom_line() +
  geom_point(size = 0.2) +
  labs(
    x = expression(df[lambda]),
    y = expression(err[lambda])
  )

err.plot
```

As expected, $\operatorname{err}$, i.e. the training error, decreases with the complexity of the model.

From example 8.7, we have that the in-sample error estimate is given as $$
\widehat{\operatorname{Err}_{\operatorname{in}}} = \operatorname{err} + \frac{2\hat\sigma^2}{n}\operatorname{trace}(\mathbf{S}) = \operatorname{err} + \frac{2\hat\sigma^2}{n}\operatorname{df}_\lambda,
$$ where we can use our estimate of $\sigma^2$ from question 5 and question 3 that the trace of $\mathbf{S}$ is the same as the effective degrees of freedom

```{r}
err_in <- err + 2 * sigma.hat.sq / n * df.path
```

This error should correct the training error by including the degrees of freeden, and thus correcting large and complex models as opposed to the training error. We plot it against the effective degrees of freedom to see if that is the case

```{r}
lambda.results <- lambda.results |>
  mutate(
    err_in = err_in
  )

ggplot(lambda.results, aes(x = df, y = err_in)) +
  geom_point(size = 0.2) +
  geom_line() +
  labs(
    x = expression(df[lambda]),
    y = expression(widehat(Err)[plain("in")])
  )
```

... And indeed it is. The in-sample error initially decreases and then it starts to increase, with the model complexity.

We can also plot both training and in-sample error together

```{r}
lambda.results |>
  pivot_longer(
    cols = c(err, err_in),
    names_to = "error_type",
    values_to = "error"
  ) |>
  mutate(
    error_type = recode(
      error_type,
      err = "Training error",
      err_in = "In-sample error"
    )
  ) |>
  ggplot(aes(x = df, y = error, colour = error_type)) +
  geom_line() +
  geom_point(size = 0.2) +
  scale_colour_manual(
    values = c(
      "Training error" = "black",
      "In-sample error" = "red"
    )
  ) +
  labs(
    x = expression(df[lambda]),
    y = "Error",
    colour = NULL
  ) 
```

Next, we go about finding the best model.

The best model is the model that minimizes the in-sample error, $\widehat{\operatorname{Err}_{\operatorname{in}}}$, and in this case, it will be over the grid we have considered.

```{r}
assessment.results <- data.frame(
  lambda = lambda.values,
  df = df.path,
  err = err,
  err.in = err_in
)

best.in.sample <- assessment.results |>
  slice_min(err.in, n = 1, with_ties = FALSE)

best.in.sample
```

We see that the best model (over the considered grid) is obtained with $\lambda_{\operatorname{opt}} \approx 0.52$ and $$
\operatorname{df}_{\operatorname{opt}} = \operatorname{df}_{\lambda_{\operatorname{opt}}} \approx 5.286, \quad \widehat{\operatorname{Err}_{\operatorname{in}}}_\text{min} \approx 316.86. 
$$

*Note that you will get different results if you consider another grid of* $\lambda$ values.

```{r}
lambda.results |>
  pivot_longer(
    cols = c(err, err_in),
    names_to = "error_type",
    values_to = "error"
  ) |>
  mutate(
    error_type = recode(
      error_type,
      err = "Training error",
      err_in = "In-sample error"
    )
  ) |>
  ggplot(aes(x = df, y = error, colour = error_type)) +
  geom_line() +
  geom_point(size = 0.2) +
  scale_colour_manual(
    values = c(
      "Training error" = "black",
      "In-sample error" = "red"
    )
  ) +
  labs(
    x = expression(df[lambda]),
    y = "Error",
    colour = NULL
  ) +
  geom_point(
    data = best.in.sample,
    aes(x = df, y = err.in),
    colour = "red",
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_vline(
    xintercept = best.in.sample$df,
    colour = "red",
    linetype = "dashed"
  )
```

We also plot the fitted values of the best in-sample model against the original unpenalized model

```{r}
ggplot(bike, aes(Temperature, Count)) +
  geom_point() +
  geom_line(
    aes(
      y = drop(X %*% beta_hat),
      colour = "Unpenalized fit"
    )
  ) +
  geom_line(
    aes(
      y = mu.hat.lambda(best.in.sample$lambda),
      colour = "Best in-sample fit"
    )
  ) +
  scale_colour_manual(
    values = c(
      "Unpenalized fit" = "black",
      "Best in-sample fit" = "blue"
    ),
    name = NULL
  )
```

#### 8. Implement cross-validation to estimate the expected generalization error. Compare graphically $\operatorname{err}$, $\widehat{\operatorname{Err}}_{\operatorname{in}}$ and $\widehat{\operatorname{Err}}$ as functions of the effective degrees of freedom.

*Note: There is a subtle difference between the following two procedures: i) splitting the data and recomputing the model matrix using `poly()` on a subset of the data; and ii) subsetting the model matrix `X` that is computed from all data. You may want to explore this difference and think about which of the two procedures is most appropriate.*

We have already compared $\operatorname{err}$ and $\widehat{\operatorname{Err}_{\operatorname{in}}}$ as functions of the effective degrees of freedom. We now also perform cross-validation.

We use squared-error-loss, $K = 10$, and $M = 20$ random fold assignments. At every data split, we use `poly` on the reduced dataset to make sure that the model matrix is orthogonal (see the note).

The procedure can be summarized as

1.  Split the observations into training and test sets.
2.  Construct the polynomial bases using the training data. Ensures that $\mathbf{X}_{\operatorname{train}}$ is orthogonal.
3.  Fit all penalized models on the training data.
4.  Use `predict()` on the training `poly` object to construct the validation basis.
5.  Evaluate squared prediction errors on the held-out observations.

Within each fold, `gamma`, `beta.hat`, `shrinkage`, and `beta.lamb` are calculated as in Questions 2 and 6.

```{r}
cv.penalized <- function(
    data,
    lamb,
    omega,
    K = 10,
    M = 20,
    degree = 6
) {
  n <- nrow(data)
  r <- length(lamb)
  loss <- vector("list", M)

  for (b in 1:M) {
    group <- sample(rep(1:K, length.out = n))
    mu.hat <- matrix(NA_real_, nrow = n, ncol = r)

    for (i in 1:K) {
      # Step 1: split into training and test.
      train <- group != i
      test <- group == i

      # Step 2: construct the polynomial basis on the training data.
      poly.train <- poly(
        data$Temperature[train],
        degree = degree
      )
      X.train <- model.matrix(~ poly.train)

      # Step 3: fit the unpenalized model.
      beta.hat <- lm.fit(
        X.train,
        data$Count[train]
      ) |>
        coefficients()

      # Fit all penalized models using the result from Question 2.
      gamma <- colSums(X.train^2)
      shrinkage <- sapply(
        lamb,
        function(lambda) {
          gamma / (gamma + lambda * omega)
        }
      )
      beta.lamb <- beta.hat * shrinkage

      # Step 4: evaluate the training basis on the test data.
      poly.test <- predict(
        poly.train,
        newdata = data$Temperature[test]
      )
      X.test <- model.matrix(~ poly.test)

      # Step 5: predict all held-out observations for all lambdas.
      mu.hat[test, ] <- X.test %*% beta.lamb
    }

    loss[[b]] <- (data$Count - mu.hat)^2
  }

  do.call(rbind, loss) |>
    colMeans()
}
```

We can perform/run the cross-validation

```{r}
set.seed(42) # For reproducibility 

cv.error <- cv.penalized(
  data = bike,
  lamb = lambda.values,
  omega = omega
)

lambda.results <- lambda.results |>
  mutate(
    cv = cv.error
  )
```

For every value of $\lambda$, `cv.error` is the average squared prediction error over all held-out observations and all $M$ repetitions.

We plot the cross-validated error as a function of the effective degrees of freedom

```{r}
cv.plot <- ggplot(
  lambda.results,
  aes(x = df, y = cv)
) +
  geom_line() +
  geom_point(size = 0.2) +
  labs(
    x = expression(df[lambda]),
    y = expression(widehat(Err)[lambda])
  )

cv.plot
```

And as for the in-sample error, we find the optimum

```{r}
best.cv <- lambda.results |>
  slice_min(cv, n = 1, with_ties = FALSE) |>
  select(lambda, df, cv)

best.cv
```

The smallest cross-validation error on the chosen grid is obtained at approximately $\lambda=`{r} round(best.cv$lambda, 3)`$, corresponding to $\operatorname{df}_\lambda=`{r} round(best.cv$df, 2)`$.

Finally, we make the overall comparison

```{r}
lambda.results |>
  pivot_longer(
    cols = c(err, err_in, cv),
    names_to = "error.type",
    values_to = "error"
  ) |>
  mutate(
    error.type = recode(
      error.type,
      err = "Training error",
      err_in = "In-sample error",
      cv = "Cross-validation error"
    )
  ) |>
  ggplot(aes(x = df, y = error, colour = error.type)) +
  geom_line() +
  scale_colour_manual(
    values = c(
      "Training error" = "black",
      "In-sample error" = "red",
      "Cross-validation error" = "blue"
    )
  ) +
  labs(
    x = expression(df[lambda]),
    y = "Squared error",
    colour = NULL
  )
```

The red dotted line marks the effective degrees of freedom selected by the estimated in-sample error, while the blue dotted line marks the effective degrees of freedom selected by cross-validation.

```{r}
plot.cv <- lambda.results |>
  pivot_longer(
    cols = c(err, err_in, cv),
    names_to = "error.type",
    values_to = "error"
  ) |>
  mutate(
    error.type = recode(
      error.type,
      err = "Training error",
      err_in = "In-sample error",
      cv = "Cross-validation error"
    )
  ) |>
  ggplot(aes(x = df, y = error, colour = error.type)) +
  geom_line() +
  geom_vline(
    xintercept = best.in.sample$df,
    colour = "red",
    linetype = "dotted"
  ) +
  geom_vline(
    xintercept = best.cv$df,
    colour = "blue",
    linetype = "dotted"
  ) +
  scale_colour_manual(
    values = c(
      "Training error" = "black",
      "In-sample error" = "red",
      "Cross-validation error" = "blue"
    )
  ) +
  labs(
    x = expression(df[lambda]),
    y = "Squared error",
    colour = NULL
  )

plot.cv
```

The training error decreases with model flexibility because it is evaluated on the same observations used for fitting. The estimated in-sample error and cross-validation error correct this optimism in different ways and therefore need not have identical minima.

The difference between the two selected models is not a contradiction. The analytic criterion estimates the in-sample error, where new outcomes are predicted at the observed predictor values, using the correction $$
\widehat{\operatorname{Err}}_{\operatorname{in}}
=
\operatorname{err}
+\frac{2\hat\sigma^2}{n}\operatorname{df}_\lambda.
$$ In contrast, 10-fold cross-validation refits the entire procedure on training samples of size $0.9n=45$ and evaluates predictions on held-out observations. It therefore estimates approximately $\operatorname{Err}(45)$ rather than the in-sample error for the model fitted to all $50$ observations. The smaller training samples make the variability of flexible fits more visible, which can make cross-validation favor stronger penalization and fewer effective degrees of freedom. Both criteria nevertheless select an intermediate model rather than either the unpenalized model or the limiting linear model.

We can also visualize the fit from the best cross-validated model

```{r}
ggplot(bike, aes(Temperature, Count)) +
  geom_point() +
  geom_line(
    aes(
      y = drop(X %*% beta_hat),
      colour = "Unpenalized fit"
    )
  ) +
  geom_line(
    aes(
      y = mu.hat.lambda(best.in.sample$lambda),
      colour = "Best in-sample fit"
    )
  ) +
    geom_line(
    aes(
      y = mu.hat.lambda(best.cv$lambda),
      colour = "Best CV fit"
    )
  ) +
  scale_colour_manual(
    values = c(
      "Unpenalized fit" = "black",
      "Best in-sample fit" = "blue",
      "Best CV fit" = "red"
    ),
    name = NULL
  )
```

Finally, we discuss the subtlety noted in the note, where we subset the model matrix instead of the data. This does not generally preserve the orthogonality of the model matrix and the result from question 2 then cannot be used.

The function below intentionally constructs `X` before cross-validation and then splits by subsetting its rows. It otherwise follows the structure of `cv.penalized()`.

```{r}
cv.penalized.matrix <- function(
    X,
    y,
    lamb,
    omega,
    K = 10,
    M = 20
) {
  n <- nrow(X)
  r <- length(lamb)
  loss <- vector("list", M)

  for (b in 1:M) {
    group <- sample(rep(1:K, length.out = n))
    mu.hat <- matrix(NA_real_, nrow = n, ncol = r)

    for (i in 1:K) {
      # Split the already constructed full-data model matrix.
      train <- group != i
      test <- group == i
      X.train <- X[train, , drop = FALSE]
      X.test <- X[test, , drop = FALSE]

      beta.hat <- lm.fit(
        X.train,
        y[train]
      ) |>
        coefficients()

      # Fit all penalized models using the result from question 2
      # even though the result does not necessarily hold in this setting
      gamma <- colSums(X^2) # perhaps colSums(X.train^2) - not affected by splits now
      shrinkage <- sapply(
        lamb, 
        function(lambda) {
            gamma / (gamma + lambda * omega)
        }
      )
      beta.lamb <- beta.hat * shrinkage

    #   # General penalized estimator: no orthogonality assumed.
    #   beta.lamb <- sapply(
    #     lamb,
    #     function(lambda) {
    #       solve(
    #         XtX + lambda * Omega,
    #         Xty
    #       )
    #     }
    #   )

      # Step 5: predict
      mu.hat[test, ] <- X.test %*% beta.lamb
    }

    loss[[b]] <- (y - mu.hat)^2
  }

  do.call(rbind, loss) |>
    colMeans()
}
```

Using the same seed ensures that this function and `cv.penalized()` use the same sequence of fold assignments.

```{r}
set.seed(42)

cv.error.matrix <- cv.penalized.matrix(
  X = X,
  y = bike$Count,
  lamb = lambda.values,
  omega = omega
)

lambda.results <- lambda.results |>
  mutate(
    cv.matrix = cv.error.matrix
  )

best.cv.matrix <- lambda.results |>
  slice_min(cv.matrix, n = 1, with_ties = FALSE) |>
  select(lambda, df, cv.matrix)

bind_rows(
  best.cv |>
    transmute(
      method = "Recompute poly() within each fold",
      lambda,
      df,
      cv.error = cv
    ),
  best.cv.matrix |>
    transmute(
      method = "Subset the full-data model matrix",
      lambda,
      df,
      cv.error = cv.matrix
    )
)
```

This second implementation is deliberately not the preferred one. First, the polynomial basis has been computed using all predictor observations before the split, so the validation predictors have influenced the preprocessing. Second, although the full matrix `X` has orthogonal columns, `X.train <- X[train, ]` generally does not. Consequently, the update $$
\hat\beta_{\lambda,j}
=
\frac{1}{1+\lambda\omega_j}\hat\beta_j
$$ is no longer justified for the row-subsetted training matrix. A mathematically correct fit for this fixed, nonorthogonal `X.train` would instead solve the general penalized normal equation for every $\lambda$; however, that would still represent a different preprocessing procedure from recomputing `poly()` within each training fold.

------------------------------------------------------------------------

## Exercise 1

Add the exercise and its solution here.

## Exercise 8.3

Let $\eta_0$ and $\eta_1$ be independent such that $\eta_1$ is distributed identically to $\eta_0 + \xi$ for some $\xi \in \mathbb{R}$. Let $F_0$ and $F_1$ denote respectively the distribution function of $\eta_0$ respectively $\eta_1$. Assume that $F_0$ is continuous. Recall that whenever the $\mathbf{E}(|f(\eta_0)|) < \infty$ we have $$
\mathbf{E}(f(\eta_0)) = \int f(x) dF_0(x),
$$ Where the right hand side is a Stieltjes integral. The above fact is sufficient to solve the exercise, but if you are unfamiliar with the Stieltjes integral and want to know more, you can consult [Wikipedia](https://en.wikipedia.org/wiki/Riemann%E2%80%93Stieltjes_integral).

### 1.

Recall that $$
\text{AUC} = \mathbf{P}(\eta_0 < \eta_1) + \frac{1}{2}\mathbf{P}(\eta_0 = \eta_1)
$$ We first show that the second term vanishes. Let $\tilde{\eta}_0$ have the same distribution as $\eta_0$ and be independent of $\eta_0$. Then, by the tower property, $$
\mathbf{P}(\eta_0 = \eta_1) = \mathbf{P}(\eta_0 = \tilde\eta_0 + \xi) = \mathbf{E}(\mathbf{P}(\eta_0 = \tilde\eta_0 + \xi|\tilde\eta_0)),
$$ and by independendence of $\eta_0$ and $\tilde \eta_0$ as well as the continuity of $F_0$, $$
\mathbf{P}(\eta_0 = \tilde\eta_0 + \xi|\tilde\eta_0) = F_0(\tilde\eta_0 + \xi) - \lim_{t \uparrow \tilde\eta_0 + \xi}F_0(t) = 0,
$$ which shows the desired. Thus, $$
\text{AUC} = \mathbf{P}(\eta_0 < \eta_1) = \mathbf{P}(\eta_0 \leq \eta_1),
$$ and applying the tower property once more shows $$
\mathbf{P}(\eta_0 \leq \eta_1) = \mathbf{E}(\mathbf{P}(\eta_0 \leq \tilde\eta_0 + \xi|\tilde\eta_0)) = \mathbf{E}(F_0(\tilde \eta + \xi)) = \int F_0(x + \xi)dF_0(x) := G(\xi).
$$ And we conclude that $$
\text{AUC} = G(\xi).
$$

### 2.

Let $\tilde{\eta}$ be as above. We can use the same argument as above to see that $$
\mathbf{P}(\eta_0 - \tilde\eta_0 \leq z) = G(z).
$$ We conclude that $G$ is the distribution function for the random variable $\eta_0 - \tilde \eta_0$

### 3.

Let now $\eta_0 \sim \mathcal{N}(\mu, \sigma^2)$. Then, $\eta_0 - \tilde{\eta}_0 \sim \mathcal{N}(0, 2\sigma^2)$. Combining this with the result from 2., we find that, $$
\text{AUC} = G(\xi) = \Phi\left(\frac{\xi}{\sqrt{2}\sigma}\right)
$$