---
title: "Week 2: Wednesday Solutions"
author: Kasper Niss, Asbjørn Risom
---

## Exercise 4.3

Show that if the structure measure $\nu = \delta_{y}$ is the Dirac-measure in $y$ then for the corresponding exponential family we have $P_{\theta} = \delta_y$ for all $\theta \in \mathbb{R}$. Show next, that if $\nu$ is not a one-point measure, then $P_{\theta}$ is not a Dirac-measure for $\theta \in I$ and conclude that its variance is strictly positive.

Assume that $\nu = \delta_{y}$. We have
$$
c(\theta) = \int e^{\theta y'} \nu(\operatorname{d} y') = \exp(\theta y) 
$$
By definition $P_{\theta}$ has density $\exp(\theta y)/\exp(\theta y) =1$ w.r.t. $\nu = \delta_{y}$ and it follows that $P_{\theta} = \delta_y$ for all $\theta \in \mathbb{R}$.

For the second claim, assume that $\nu$ is not a one-point measure and for contradiction that there exists a $\theta \in I$ and a $y \in \mathbb{R}$ such that $P_{\theta} = \delta_{y}$. Then
$$
\frac{\operatorname{d}P_{\theta}}{\operatorname{d} \nu} (y) = \frac{1}{c(\theta)}\exp(\theta y )
$$
And 
$$
1 = P_{\theta}(\{y\}) = \frac{1}{c(\theta)} \int_{\{y\}} \exp(\theta y') \nu(dy')
$$
But this implies that
$$
\int_{\{y\}} \exp(\theta y') \nu(dy') = \int \exp(\theta y') \nu(dy') \implies \int_{\mathbb{R} \setminus \{y\}} \exp(\theta y') \nu(dy') = 0 
$$
If we can show that this implies that $\nu( \mathbb{R} \setminus \{y\}) = 0$ we are done, as $\nu$ is not the zero measure, we would have $\nu(\{y\}) > 0$, which implies that $\nu$ is a one-point measure, a contradiction. However, this is a standard measure theory result, let $(\mathcal{X}, \mathcal{F},\mu)$ be a measure space. Let $A \in \mathcal{F}$ and suppose that $f > 0$ on $A$ and 
$$
\int_A f(x) \mu(\operatorname{d}x) = 0
$$
Set $A_n = A \cap (f \geq 1/n)$, such that $\cup_{n} A_n = A$. Then 
$$
0=\int_{A} f(x) \mu(\operatorname{d}x) = \int_{A_n} f(x) \mu(\operatorname{d}x) \geq \frac{1}{n}\nu(A_n),
$$
Such that $\nu(A_n) = 0$ for all $n$. In particular $\nu(A) \leq \sum_{n=1}^{\infty}\nu(A_n) = 0$.

## Exercise 4.7

Let $\nu = m_{[-1,1]}$ be the Lebesgue measure restricted to [-1,1]$.

#### 1.

Since $e^{\theta y }$ is bounded over $[-1,1]$ for any $\theta \in \mathbb{R}$, we have $c(\theta) < \infty$ for all $\theta \in \mathbb{R}$, from which we conclude that $I = \mathbb{R}$. We compute the unit cumulant function $\kappa$.
$$
c(\theta) = \int e^{\theta y} \nu(\operatorname{d}y) = \int_{-1}^1 e^{\theta y} \operatorname{d}y = \frac{1}{\theta}(e^{\theta}-e^{-\theta}) = 2\frac{\sinh(\theta)}{\theta}.
$$
Which gives us the unit cumulant 
$$
\kappa(\theta) = \log(c(\theta)) = \log(2) + \log(\sinh(\theta)) - \log(\theta)
$$.

We plot the density w.r.t. Lebesgue measure for $\theta = 1$ and $\theta = -1$

```{r}
#| message: false
#| warning: false
library(ggplot2)
library(dplyr)
c_theta <- function(theta){
  2*sinh(theta)/theta
}
dens <- function(theta, y){
  exp(theta*y)/c_theta(theta)
}
ys <- seq(-1,1, by =0.001)
dens1 <- dens(1,ys)
densneg1 <- dens(-1,ys)
df_plot <- tibble(y=rep(ys,2), dens = c(dens1,densneg1), theta = c(rep(1,length(ys)),rep(-1,length(ys))))
ggplot(df_plot, aes(x = y, y = dens, color = as.factor(theta))) + geom_line() + 
  xlab("y") + ylab("Density") + 
  theme_bw()

```

#### 2.

Let $Y \sim P_{\theta}$ where
$$
\frac{\operatorname{d}P_{\theta}}{\operatorname{d}\nu} = e^{\theta y - \kappa(\theta)}
$$
We compute the mean value map by differentiating the unit cumulant (Theorem 4.1). Recalling that the derivative of $\sinh$ is $\cosh$ we get,
$$
\mathbf{E}(Y) = \kappa'(\theta) = \frac{\cosh(\theta)}{\sinh(\theta)} - \frac{1}{\theta},
$$
and recalling that $\cosh$ has derivative $-\sinh$ we find that the variance is, for $\theta \neq 0$
$$
\mathbf{V}(Y) = \kappa''(\theta)= \frac{1}{\theta^2} - \frac{\sinh(\theta)^2 - \cosh(\theta)^2 }{\sinh(\theta)^2} = \frac{1}{\theta^2} - \frac{1 }{\sinh(\theta)^2}
$$
We plot the variance function $\mathcal{V}(\mu)$ for varying $\mu$.


```{r}
mean_value <- function(theta){
  1/cosh(theta) - 1/theta
} 
var_fun <- function(theta){
  1/theta^2 - 1/sinh(theta)^2
}
thetas <- c(seq(-2,-0.01, by = 0.001),seq(0.01,2, by =0.001 ))
mu <-  mean_value(thetas)
var_mu <- var_fun(thetas)
ggplot(data = tibble(mu = mu, var_mu = var_mu), aes(mu,var_mu)) + geom_line() + theme_bw()

```


## Exercise 5.2

Let for $\eta, \rho > 0$ 
$$
p_{\eta}(k) = G(k,\rho) \left( \frac{1}{\rho + \eta} \right)^{\rho} \left( \frac{\eta}{\rho + \eta} \right)^{k}
$$
Fix $\rho$ throughout.

#### 1.

We want to show that $p_{\eta}$ is an exponential family. Note
$$
\left( \frac{\eta}{\rho + \eta} \right)^{k} = e^{k \log \frac{\eta}{\rho + \eta}}
$$
Set $\theta = \log \frac{\eta}{\rho + \eta}$ such that
$$
e^{\theta}(\rho + \eta) = \eta \Leftrightarrow \eta(1-e^{\theta}) = \rho e^{\theta} \Leftrightarrow \eta = \rho \frac{e^{\theta}}{1-e^\theta},
$$
We obtain
$$
\rho + \eta = \rho \left(1 + \frac{e^{\theta}}{1-e^\theta} \right) = \rho \frac{1}{1-e^\theta}.
$$
In this parameterization, we can thus write
$$
p_{\eta}(k) = G(k,p) \left( \frac{1-e^{\theta}}{\rho} \right)^\rho \exp(\theta y )
$$
If we choose $\nu = G(k,p) \rho^{-\rho} \cdot \tau$, with $\tau$ the counting measure on $\mathbb{N}$ as our structure measue we obtain an exponential family. The normalizing constant is
$$
c(\theta) = (1-e^{\theta})^{-\rho},
$$
and the unit cumulant is
$$
\kappa(\theta) = -\rho \log(1-e^\theta)
$$

#### 2.

We compute the mean value function by applying theorem 4.1,
$$
\kappa'(\theta) = -\rho \frac{-e^{\theta}}{1-e^{\theta}} = \rho \frac{e^{\theta}}{1-e^{\theta}} = \eta = \mu(\eta).
$$
We compute the variance function by taking the second derivative,
$$
\kappa''(\theta) = \rho \frac{e^\theta(1-e^{\theta})+e^{2\theta}}{(1-e^{\theta})^2} = \rho \frac{e^{\theta}}{(1-e^{\theta})^2} = \eta \frac{1}{1-e^\theta}
$$
$$
1-e^{\theta} = 1-\frac{\eta}{\rho + \eta} = \frac{\rho}{\rho + \eta}
$$
So,
$$
\kappa''(\theta) = \eta \frac{\rho + \eta}{\rho} = \eta + \frac{\eta^2}{\rho} = \mathcal{V}(\mu(\eta))
$$

#### 3.

We implement the Fisher scoring algorithm and implement for various values of $\rho$:

```{r}
Xi <- c(5,10,15,20,25,30,35,40,45,50)
Yi <- c(6,11,16,38,82,22,108,61,55,64)

#Designmatrix 
X <- matrix(c(rep(1,10),Xi),ncol=2)

update <- function(beta, rho){
  eta <- X %*% beta
  mu <- eta
  Z <- Yi
  W <- c(rho / (eta * (eta + rho)))
  return(lm(Z ~ Xi,weights=W)$coefficients)
}

IWLS <- function(beta.init, rho = 1, eps = 10^-6, numIter = 10){
  iter <- 0
  beta.old <- beta.init
  for(i in 1:numIter){
    beta.new <- update(beta.old, rho)
    if(dist(rbind(beta.old, beta.new)) < eps){
      iter <- i
      break
    }
    beta.old <- beta.new
  }
  return(list(beta = beta.new, iter = iter))
}


IWLS(beta.init = c(1, 1),rho = 1)
IWLS(beta.init = c(1, 1),rho = 10)
IWLS(beta.init = c(1, 1),rho = 0.1)
```
We see that the coefficients are not very sensitive to the choice of $\rho$.

#### 4.

We have that the pearson residuals are given by:
$$ 
\frac{Y_i - \mu(\hat{\eta_i})}{\sqrt{ \mathcal{V}(\mu(\hat{\eta_i}))}} = \frac{Y_i - \hat{\eta_i}}{\sqrt{\hat{\eta_i} + \frac{\hat{\eta_i}^2}{\rho}}}
$$
```{r}
# Find MLE using our algorithm for rho = 1
rho <- 1
betahat <- IWLS(beta.init = c(1, 1), rho = rho)$beta
etahat <- X %*% betahat


# Pearson residuals 
resid <- (Yi - etahat)/(sqrt(etahat + etahat^2/rho))

# The fitted values (the etas) against the residuals: 
df_Pearson_resid <- data.frame(.fitted = etahat, .resid = resid)
ggplot(aes(.fitted,.resid), data = df_Pearson_resid) +
	geom_point() +
  geom_smooth() + 
  xlab("Fitted values") + 
  ylab("Pearson residuals") + 
  theme_bw()
```
The Pearson $\chi^2$ statistic is
```{r}
sum(resid^2)
```

#### 5.

We implement a function that computes the $\chi^2$ statistic as a function of $\rho$. This is just a matter of generalizing the above computation. Recall equation (5.6) which gives the following formula for an estimator of the dispersion parameter in a generalized linear model.
$$
\hat{\psi} = \frac{1}{n-p}\chi^2
$$
In our case, we have an exponential family, i.e. unit dispersion, and as $n-p = 8$ our case, we therefore expect $8 =\chi^2$. 
```{r}
chisq <- function(rho) 
  {
  betahat <- IWLS(beta.init = c(1, 1), rho)$beta
  etahat <- c(X %*% betahat)
  # Pearson residuals 
  resid <- (Yi - etahat)/(sqrt(etahat + etahat^2/rho))
  sum(resid^2)
}

rhoseq <- seq(0.1, 5, 0.01)
chiseq <- sapply(X = rhoseq, FUN = chisq)

df_Chisq <- data.frame(rho = rhoseq, chisq = chiseq)

ggplot(data = df_Chisq, aes(x = rho, y = chisq)) +
  geom_line() +
  geom_hline(yintercept = 8, linetype = "dashed", color = "red") +
  xlab(expression(rho)) +
  ylab(expression(chi^2))+ 
  theme_bw()
```
We see that $\chi^2 = 8$ approximately when $\rho = 4$. Data seems to indicate that $\rho = 4$ is a reasonable choice.

#### 6.

We use equation (5.7) to compute confidence intervals for $\beta_1$, where we obtain $\beta_1$ using MLE under $\rho =1$, and base our confidence intervals on the estimate of the dispersion parameter from above, when $\rho = 1$ respectivly 4.

```{r}
#Find MLE using our algorithm for rho = 1
betahat <- IWLS(c(1,1), 1)$beta
etahat <- X %*% betahat

# Computing the two different values of the dispersion parameter (equation 6.5)
psihat1 <- 1/8*chisq(rho = 1)
psihat4 <- 1/8*chisq(rho = 4)

# Computing Weight matrix in both cases
What1 <- diag( as.vector( 1 / (etahat*(1 + etahat)) ))
What4 <- diag( as.vector( 4 / (etahat*(4 + etahat)) ))

#the standard errors
sd1hat <- sqrt(psihat1 %*% solve(t(X )%*% What1 %*% X)[2,2])
sd4hat <- sqrt(psihat4 %*% solve(t(X )%*% What4 %*% X)[2,2])

#The confidence intervals 
z <- qnorm(0.975)
betahat[2] + c(-z*sd4hat, z*sd4hat)
betahat[2] + c(-z*sd1hat, z*sd1hat)
```

We see that the results are almost identical.
